---
title: "Tutorial 4: Phylogenetic Validation and Comparative Analyses"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 4: Phylogenetic Validation and Comparative Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The final stage of the `PhyloCactus` framework evaluates topological stability and comparative congruence between the inferred evolutionary hypothesis and external reference backbones.

Although maximum-likelihood inference provides a statistically evaluated estimate of evolutionary relationships, the resulting supermatrix topology represents a single analytical hypothesis. In rapid plant radiations such as **Cactaceae**, topological recovery can be influenced by taxon sampling, locus selection, substitution model parametrization, and analytical strategy.

This module quantifies topological agreement between the focal maximum-likelihood supermatrix tree generated by `PhyloCactus` and previously published reference phylogenetic hypotheses for Cactaceae, of which only the nuclear target-capture dataset of de Vos *et al.* (2025) and Zuntini *et al.* (2024) are phylogenomic in scale (hundreds of nuclear genes); the Amaral *et al.* (2022) and Thompson *et al.* (2024) frameworks are macroevolutionary and spatial-diversity syntheses rather than de novo phylogenomic reconstructions. Validation employs quantitative tree comparison metrics, including Robinson-Foulds (RF) distances, normalized topological distances, and multidimensional scaling (MDS) tree-space projections. These approaches establish an objective framework to identify congruent evolutionary clades, pinpoint conflicting topologies, and evaluate the placement of major cactus lineages across these alternative reference datasets.

## Complete Pipeline Execution Workflow

### Module 13: Sub-trees and Phylogenetic Validation

#### Sourcing External Trees and `ASTRAL-III`

Reference phylogenetic hypotheses and external analytical binaries can be supplied by the user via environment variables or local paths. To ensure immediate, reproducible execution without mandatory external downloads, reference topologies (Amaral *et al*. 2022; Thompson *et al*. 2024; de Vos *et al*. 2025; Zuntini *et al*. 2024) are distributed directly in `PhyloCactus` under `inst/extdata/`.

The validation protocol compares the focal supermatrix topology against four representative reference frameworks: the spatial phylodiversity tree of Amaral *et al.* (2022; *Biological Conservation*, 273), the macroevolutionary chronogram of Thompson *et al.* (2024; *Nature Communications*, 15), the phylogenomic nuclear target-capture dataset of de Vos *et al.* (2025; *Plant Systematics and Evolution*, 311), and the Kew Tree of Life backbone of Zuntini *et al.* (2024; *Nature*, 629).

#### Gene Tree Summaries with `ASTRAL-III` under Multispecies Coalescence

In multilocus and target-capture datasets, individual gene trees frequently display topological discordance. Rather than reflecting analytical noise, this discordance stems from primary biological processes, including Incomplete Lineage Sorting (ILS), reticulate evolution, ancient hybridization, and plastid capture. Because gene coalescence occurs within the bounds of ancestral species lineages, concatenated supermatrix approaches can occasionally suffer from inconsistency under high ILS.

To explicitly account for biological discordance, `PhyloCactus` incorporates the summary species-tree estimator `ASTRAL-III`. Operating under the multispecies coalescent (MSC) framework, `ASTRAL-III` infers a species tree topology from a collection of unrooted gene trees, minimizing deep coalescent events. This provides a coalescent-based reference hypothesis to compare directly against the concatenated maximum-likelihood reconstruction.

The `ASTRAL-III` Java executable (`.jar`) is downloaded separately by the user before running the coalescent summary workflow.

#### Using `.Renviron` to Manage Paths Safely

When dealing with personal file paths (to software like `ASTRAL-III` or locally downloaded trees), it is highly recommended to use the `.Renviron` file. This avoids exposing your local machine paths on the internet or in your shared code.

```{r, eval = FALSE}
# Add these lines to your ~/.Renviron file (e.g., using usethis::edit_r_environ())
# ASTRAL_PATH="/path/to/astral.5.7.8.jar"
# DEVOS_GENETREES_PATH="/path/to/QC.bestTreeCollapsed.trees"
# DEVOS_METADATA_PATH="/path/to/606_2025_1948_MOESM1_ESM.csv"

# Then access them safely in R:
astral_jar <- Sys.getenv("ASTRAL_PATH")
gene_trees_file <- Sys.getenv("DEVOS_GENETREES_PATH")
metadata_file <- Sys.getenv("DEVOS_METADATA_PATH")
```

#### Standardizing Metadata and Running `ASTRAL-III`

*Context: Why use `ASTRAL-III`?* The study by de Vos *et al*. (2025) is a phylogenomic study utilizing the *Angiosperms353* target capture kit. Because their analysis was centered primarily at the genus level, it provides an excellent framework for testing topological congruence against our supermatrix approach. However, because the authors provided individual gene trees rather than the final coalescent species tree, we must infer the species tree using `ASTRAL-III` to properly account for gene tree discordance (e.g., resulting from incomplete lineage sorting) prior to our comparative analyses.

Using the gene trees and metadata (e.g., from de Vos *et al.* 2025), you can map sample names, export renamed gene trees, and infer the coalescent species tree. The resulting tree is stored directly in the tutorial validation directory:

```{r}
# Target output paths in the validation directory
val_dir <- file.path(tempdir(), "10_Validation")
dir.create(val_dir, recursive = TRUE, showWarnings = FALSE)
output_tree <- file.path(val_dir, "QC.Species_tree_astral.tree")
renamed_gene_trees_file <- file.path(val_dir, "QC.Species_tree_astral_rename.tree")

# Flag controlling live ASTRAL-III execution. Default is FALSE to ensure fast,
# reproducible compilation across all build environments.
run_astral_locally <- FALSE

astral_jar       <- Sys.getenv("ASTRAL_PATH")
devos_gene_trees <- Sys.getenv("DEVOS_GENETREES_PATH")
devos_metadata   <- Sys.getenv("DEVOS_METADATA_PATH")

java_bin <- Sys.which("java")
if (!nzchar(java_bin) && file.exists("/opt/homebrew/opt/openjdk/bin/java")) {
  java_bin <- "/opt/homebrew/opt/openjdk/bin/java"
}

if (run_astral_locally && nzchar(astral_jar) && file.exists(astral_jar) && 
    nzchar(devos_gene_trees) && file.exists(devos_gene_trees) && 
    nzchar(devos_metadata) && file.exists(devos_metadata) && 
    nzchar(java_bin)) {
  
  meta <- read.csv(devos_metadata, stringsAsFactors = FALSE, check.names = FALSE)
  meta$species_name <- gsub(" ", "_", meta$`Scientific name`)
  meta$Sample_from_tree <- stringr::str_extract(meta$`Name in tree`, "P[0-9]+")
  
  map_sample <- setNames(meta$species_name, meta$Sample_from_tree)
  map_srr    <- setNames(meta$species_name, meta$`ENA run acc.`)
  
  gene_trees <- ape::read.tree(devos_gene_trees)
  
  rename_tips <- function(tree) {
    tips <- tree$tip.label
    paftol_to_sample <- function(x) {
      num <- stringr::str_extract(x, "[0-9]+")
      paste0("P", stringr::str_sub(num, -5))
    }
    sample_ids <- ifelse(stringr::str_detect(tips, "PAFTOL"), paftol_to_sample(tips), NA)
    species_vec <- ifelse(stringr::str_detect(tips, "PAFTOL"), map_sample[sample_ids], map_srr[tips])
    species_vec[is.na(species_vec)] <- tips[is.na(species_vec)]
    tree$tip.label <- species_vec
    return(tree)
  }
  
  gene_trees <- lapply(gene_trees, rename_tips)
  class(gene_trees) <- "multiPhylo"
  ape::write.tree(gene_trees, file = renamed_gene_trees_file)
  
  cmd <- paste(shQuote(java_bin), "-jar", shQuote(astral_jar), 
               "-i", shQuote(renamed_gene_trees_file), 
               "-o", shQuote(output_tree), "2>/dev/null")
  system(cmd)
} else {
  # Deploy precalculated de Vos et al. (2025) reference tree to ensure
  # the target tree file is always generated in the validation directory
  ref_devos <- system.file("extdata", "reference_devos_2025_astral.tree", package = "PhyloCactus")
  file.copy(ref_devos, output_tree, overwrite = TRUE)
}
```

### Evaluating Topological Congruence Across Alternative Hypotheses

Once external phylogenetic hypotheses have been downloaded, and new `ASTRAL-III` species trees inferred, we evaluate topological congruence against our focal maximum likelihood tree (the `treePL` chronogram) and external reference trees.

Topological congruence is quantified using Robinson-Foulds (RF) distances and related tree comparison metrics. These analyses measure the number of shared and conflicting bipartitions among alternative phylogenetic hypotheses, allowing researchers to identify the degree of agreement among datasets generated under different taxon sampling schemes, molecular evidence, and inference frameworks.

For visualization of relationships among alternative hypotheses, normalized RF distances are summarized using multidimensional scaling (MDS). This ordination approach represents phylogenetic hypotheses in a reduced dimensional space, where closer positions indicate greater topological similarity and larger distances indicate stronger disagreement.

**Rooting note:** `validate_phylogenies()` roots every tree on *Leuenbergeria*, which differs deliberately from the outgroup rooting (Talinaceae and Portulacaceae) used in Module 10. Comparison requires a shared taxon set, so the function first restricts every tree to **Cactaceae**, and that filter removes the Talinaceae, Portulacaceae, and Anacampserotaceae terminals that carry the root in the dating pipeline. *Leuenbergeria* is the earliest-diverging lineage retained, which makes it the closest available approximation to the root position the full tree imposes. The genus is exposed as `rooting_genus` for datasets sampled differently, and trees lacking it are left unrooted and flagged in `SUPP_rooting_summary`. Robinson-Foulds distances are computed on unrooted bipartitions, so this choice affects the tree-space projection and the reported rooting summary rather than the distances themselves.

**Standardized benchmark comparison:** To conduct an objective, standardized, and honest comparative validation across alternative evolutionary hypotheses, all comparative trees (`FocalTree`, `Zuntini`, `Thompson`, `Amaral`, and `deVos`) are loaded directly from the validated benchmark dataset in `inst/extdata/`. This approach ensures that every tree comparison uses an identical taxonomic crosswalk without introducing methodological discrepancies between partially executed local runs and published reference backbones.

```{r, fig.width=11.69, fig.height=8.27, warning=FALSE, message=FALSE}
library(PhyloCactus)

# List of collected trees: loaded from the standardized extdata benchmark suite
tree_paths <- list(
  FocalTree   = system.file("extdata", "phylocactus_chronogram_hpd.tree", package = "PhyloCactus"),
  Zuntini     = system.file("extdata", "reference_zuntini_2024.tree", package = "PhyloCactus"),
  Thompson    = system.file("extdata", "reference_thompson_2024.tree", package = "PhyloCactus"),
  Amaral      = system.file("extdata", "reference_amaral_2022.tree", package = "PhyloCactus"),
  deVos       = system.file("extdata", "reference_devos_2025_astral.tree", package = "PhyloCactus")
)

# Filter missing trees to allow the vignette to build even if you haven't uploaded them yet
existing_trees <- tree_paths[sapply(tree_paths, function(x) file.exists(x) && nzchar(x))]

plot_to_show <- NULL
table_to_show <- NULL

if(length(existing_trees) > 1) {
  checklist_file   <- system.file("extdata", "CactaceaeFullList_2026_07_01_Beatriz_Merino.xlsx", package = "PhyloCactus")
  constraints_file <- system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus")
  
  # The validate_phylogenies wrapper handles tip harmonization, Robinson-Foulds/DPI distance
  # calculations, and tree-space projections automatically.
  validation_outputs <- validate_phylogenies(
    trees_mapping_list = existing_trees,
    checklist_csv = checklist_file,
    constraints_map = constraints_file,
    out_dir = file.path(tempdir(), "10_Validation")
  )
  
  plot_to_show <- validation_outputs$plot
  
  dist_table <- read.csv(file.path(validation_outputs$tables, "TABLE_pairwise_tree_distances_main.csv"))
  table_to_show <- head(dist_table)
} else {
  message("Please upload the reference trees to inst/extdata/ to run the full validation.")
}

plot_to_show
```

```{r, echo=FALSE}
if (!is.null(table_to_show)) {
  knitr::kable(table_to_show, caption = "Pairwise tree distances")
}
```

The resulting `TABLE_pairwise_tree_distances_main.csv` and the multidimensional tree-space projections (`FIG_tree_space_main.pdf`) provide a quantitative framework for assessing topological congruence among alternative phylogenetic hypotheses. Together, these outputs allow researchers to identify regions of agreement and disagreement across studies, evaluate the robustness of major clades, and determine whether differences arise from taxon sampling, molecular marker selection, or phylogenetic inference methodology. This comparative validation provides an objective measure of confidence in the inferred evolutionary relationships before downstream macroevolutionary, biogeographic, or comparative analyses.

## Conclusion

This concludes the core `PhyloCactus` phylogenetic inference pipeline. Across the preceding tutorials, you have assembled a curated multilocus supermatrix from public sequence repositories, inferred a maximum-likelihood phylogeny, estimated divergence times using penalized likelihood with temporal bootstrap confidence intervals, integrated conservation metadata from the IUCN Red List, generated publication-ready visualizations, and quantitatively validated your phylogenetic hypothesis against alternative published backbones.

The resulting datasets (including aligned supermatrices, maximum likelihood trees, time calibrated chronograms, species metadata tables, conservation summaries, and comparative validation metrics) provide a reproducible foundation for downstream analyses in macroevolution, historical biogeography, conservation biology, community ecology, and comparative evolutionary research. By combining automated data acquisition, rigorous quality control, standardized phylogenetic inference, and transparent analytical workflows, `PhyloCactus` enables researchers to construct large scale evolutionary datasets that are readily reproducible, extensible, and suitable for publication quality comparative studies.

The next stage of the `PhyloCactus` workflow is intended to evaluate the molecular diagnostic resolution of individual markers. That section is under construction and ships no implementation: its analytical strategy has not been defined, and the earlier routines were withdrawn rather than left in place as working code whose output could be mistaken for a validated result. The following tutorial is a manifest setting out the scope of the section, the design decisions already taken, the questions that remain open, and the criteria for reopening it.

[Continue to Tutorial 5: DNA Barcoding, Section Manifest](https://beeamerino.github.io/PhyloCactus/articles/tutorial-5-cactus-phylogeny-barcoding.html)

## References

- Amaral *et al*. 2022. Spatial patterns of evolutionary diversity in Cactaceae show low ecological representation within protected areas. *Biological Conservation*, 273, 109677. <https://doi.org/10.1016/j.biocon.2022.109677>
- de Vos, J. M., Eggli, U., Nyffeler, R., Larridon, I., McGinnie, C., Epitawalage, N., Maurin, O., Forest, F., & Baker, W. J. 2025. Phylogenomics and classification of Cactaceae based on hundreds of nuclear genes. *Plant Systematics and Evolution*, 311(5), 28. <https://doi.org/10.1007/s00606-025-01948-z>
- Thompson *et al*. 2024. Identifying the multiple drivers of cactus diversification. *Nature Communications*, 15(1), 7114. <https://doi.org/10.1038/s41467-024-51666-2>
- Zuntini, A. R., Frankel, F. L., Pokorny, L., Forest, F., Baker, W. J., et al. 2024. Phylogenomics and the rise of the angiosperms. *Nature*, 629, 843-850. <https://doi.org/10.1038/s41586-024-07324-0>

