---
title: "Tutorial 6: Core Pipeline Functions and Methodological Dictionary"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 6: Core Pipeline Functions and Methodological Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This page lists the main functions of the `PhyloCactus` workflow, grouped by stage, with the method each one implements and its primary reference.
Arguments and return values are documented in the manual pages and in the `pkgdown` reference index.

## Functions by Stage

### Sequence Assembly

| Function | Description | Reference |
|:---|:---|:---|
| `assemble_ingroup_phylotar()` | Retrieves sequence clusters for the ingroup (Cactaceae, NCBI Taxonomy ID 3593) from GenBank with `phylotaR`. Clusters are formed by sequence similarity, not by GenBank locus annotations; gene-name synonyms are mapped with `genes_map.csv`, clusters with more than `min_species` species are retained, and a curated list of excluded accessions is applied. | Bennett et al. (2018) |
| `assemble_outgroup_phylotar()` | Retrieves sequence clusters for the outgroup genera of Anacampserotaceae, Portulacaceae and Talinaceae, for the loci selected in the ingroup. | Bennett et al. (2018) |
| `fetch_genbank_metadata()` | Retrieves the organism name and definition line of each GenBank accession through `ape::read.GenBank()`, in batches with retries, and caches them. | Paradis & Schliep (2019) |

### Alignment and Quality Control

| Function | Description | Reference |
|:---|:---|:---|
| `run_alignment_pipeline()` | Aligns each locus with `MAFFT` and masks poorly aligned columns with `DECIPHER::MaskAlignment()`. | Katoh & Standley (2013); Wright (2024) |
| `run_marker_screening()` | Screens each locus alignment for coverage, length and substitution saturation, by regressing uncorrected p-distance against the gamma-corrected K80 distance. The slope threshold below which a locus is rejected is a `PhyloCactus` criterion, not a published saturation test; the references cover the distance model only. | Kimura (1980); Jin & Nei (1990) |
| `clean_taxonomic_names()` | Keeps the sequences whose name is in the checklist of Cactaceae, ignoring differences in spaces, hyphens and underscores, and renames them to the spelling of the checklist. | Korotkova et al. (2021) |
| `run_joint_realignment()` | Realigns each locus with `MAFFT` after the ingroup and outgroup sequences are combined. | Katoh & Standley (2013) |

### Supermatrix

| Function | Description | Reference |
|:---|:---|:---|
| `run_concatenation_pipeline()` | Concatenates the locus alignments into a supermatrix and writes partition maps in RAxML format (read by `RAxML-NG`, `ModelTest-NG` and `IQ-TREE`) and NEXUS format, with configuration files for `PartitionFinder2` and `MrBayes`. | Not applicable |

### Model Selection and Maximum-Likelihood Inference

| Function | Description | Reference |
|:---|:---|:---|
| `preprocess_partitions()` | Checks the supermatrix and partition map with `RAxML-NG` before model selection. | Kozlov et al. (2019) |
| `run_modeltest_ng()` | Selects a substitution model for each partition with `ModelTest-NG` under the AICc. | Darriba et al. (2020) |
| `build_constraint_scaffold()` | Builds a multifurcating constraint tree that enforces the monophyly of subfamilies and major clades of Cactaceae established in published phylogenies. | Guerrero et al. (2019) |
| `calculate_ml_tree()` | Estimates the maximum-likelihood tree with `RAxML-NG` under the partition models and the constraint tree, from 25 random and 25 parsimony starting trees by default. | Kozlov et al. (2019) |
| `map_branch_supports()` | Maps Felsenstein bootstrap proportions (FBP) or transfer bootstrap expectation (TBE) from the bootstrap replicates onto the maximum-likelihood tree. | Felsenstein (1985); Lemoine et al. (2018) |

### Divergence Time Estimation

| Function | Description | Reference |
|:---|:---|:---|
| `rescale_tree()` | Multiplies all branch lengths by a constant before dating with `treePL`. The factor changes the scale of the rate parameters and the numerical conditioning of the optimization; it does not change which branches `treePL` treats as too short. | Smith & O'Meara (2012) |
| `run_treePL_direct()` | Runs the `treePL` binary on a configuration file, after checking that the tree is rooted. | Smith & O'Meara (2012) |
| `run_treePL_cv()` | Selects the rate-smoothing parameter by cross-validation, classifies the shape of the cross-validation curve, and dates the tree. | Sanderson (2002); Maurin (2020) |
| `automate_treePL()` | Runs cross-validation, smoothing selection and dating on the maximum-likelihood tree and on each bootstrap replicate. | Smith & O'Meara (2012); Maurin (2020) |
| `send_run_notification()` | Sends an email when a long analysis ends, reporting status, timings and output paths. A failure to send does not stop the analysis. | `blastula` R package, <https://github.com/rstudio/blastula> |

### Comparison, Figures and IUCN Data

| Function | Description | Reference |
|:---|:---|:---|
| `validate_phylogenies()` | Compares the focal tree with published trees using the Robinson-Foulds and Different Phylogenetic Information distances, and projects the distances into tree space by multidimensional scaling. | Robinson & Foulds (1981); Smith (2020) |
| `integrate_publication_tree()` | Maps support values onto the chronogram and the maximum-likelihood tree and collapses nodes with FBP below 70 into polytomies for the publication figures. | Hillis & Bull (1993) |
| `get_iucn_data()` | Retrieves the IUCN Red List assessment of a species through `rredlist`: category, criteria, population trend, countries of occurrence, endemism, habitats and threats. | IUCN (2026) |
| `scale_fill_iucn()` | `ggplot2` fill scale with the standard IUCN Red List category colors. | IUCN (2026) |

### Help Pages

```{r, eval = FALSE}
library(PhyloCactus)

?run_alignment_pipeline
?calculate_ml_tree
?automate_treePL
```

The functions used for checks within the tutorials, such as `check_calibration_consistency()`, `report_bound_adherence()`, `classify_constrained_nodes()`, `audit_calibration_support()` and `report_smoothing_sensitivity()`, are documented in their manual pages and in the tutorials where they are used.

## References

- Bennett, D. J., Hettling, H., Silvestro, D., Zizka, A., Bacon, C. D., Faurby, S., ... & Antonelli, A. 2018. phylotaR: An automated pipeline for retrieving orthologous DNA sequences from GenBank in R. *Life*, 8(2), 20. <https://doi.org/10.3390/life8020020>
- Darriba, D., Posada, D., Kozlov, A. M., Stamatakis, A., Morel, B., & Flouri, T. 2020. ModelTest-NG: a new and scalable tool for the selection of DNA and protein evolutionary models. *Molecular Biology and Evolution*, 37(1), 291-294. <https://doi.org/10.1093/molbev/msz189>
- Felsenstein, J. 1985. Confidence limits on phylogenies: an approach using the bootstrap. *Evolution*, 39(4), 783-791. <https://doi.org/10.1111/j.1558-5646.1985.tb00420.x>
- Guerrero, P. C., Majure, L. C., Cornejo-Romero, A., & Hernández-Hernández, T. 2019. Phylogenetic Relationships and Evolutionary Trends in the Cactus Family. *Journal of Heredity*, 110(1), 4-21. <https://doi.org/10.1093/jhered/esy064>
- Hillis, D. M., & Bull, J. J. 1993. An empirical test of bootstrapping as a method for assessing confidence in phylogenetic analysis. *Systematic Biology*, 42(2), 182-192. <https://doi.org/10.1093/sysbio/42.2.182>
- IUCN. 2026. *The IUCN Red List of Threatened Species*. Version 2026-1. Accessed 2026-09-19. <https://www.iucnredlist.org>
- Jin, L., & Nei, M. 1990. Limitations of the evolutionary parsimony method of phylogenetic analysis. *Molecular Biology and Evolution*, 7(1), 82-102. <https://doi.org/10.1093/oxfordjournals.molbev.a040588>
- Katoh, K., & Standley, D. M. 2013. MAFFT multiple sequence alignment software version 7: Improvements in performance and usability. *Molecular Biology and Evolution*, 30(4), 772-780. <https://doi.org/10.1093/molbev/mst010>
- Kimura, M. 1980. A simple method for estimating evolutionary rates of base substitutions through comparative studies of nucleotide sequences. *Journal of Molecular Evolution*, 16(2), 111-120. <https://doi.org/10.1007/BF01731581>
- Korotkova, N., Aquino, D., Arias, S., Eggli, U., Franck, A., Gómez-Hinostrosa, C., Guerrero, P. C., Hernández, H. M., Kohlbecker, A., Köhler, M., Luther, K., Majure, L. C., Müller, A., Metzing, D., Nyffeler, R., Sánchez, D., Schlumpberger, B. O., & Berendsohn, W. G. 2021. Cactaceae at Caryophyllales.org, a dynamic online species-level taxonomic backbone for the family. *Willdenowia*, 51(2), 251-270. <https://doi.org/10.3372/wi.51.51208>
- Kozlov, A. M., Darriba, D., Flouri, T., Morel, B., & Stamatakis, A. 2019. RAxML-NG: a fast, scalable and user-friendly tool for maximum likelihood phylogenetic inference. *Bioinformatics*, 35(21), 4453-4455. <https://doi.org/10.1093/bioinformatics/btz305>
- Lemoine, F., Domelevo Entfellner, J. B., Wilkinson, E., Correia, D., Dávila Felipe, M., De Oliveira, T., & Gascuel, O. 2018. Renewing Felsenstein's phylogenetic bootstrap in the era of big data. *Nature*, 556(7702), 452-456. <https://doi.org/10.1038/s41586-018-0043-0>
- Maurin, K. J. L. 2020. An empirical guide for producing a dated phylogeny with treePL in a maximum likelihood framework. *arXiv*:2008.07054. <https://doi.org/10.48550/arXiv.2008.07054>
- Paradis, E., & Schliep, K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. *Bioinformatics*, 35(3), 526-528. <https://doi.org/10.1093/bioinformatics/bty633>
- Robinson, D. F., & Foulds, L. R. 1981. Comparison of phylogenetic trees. *Mathematical Biosciences*, 53(1-2), 131-147. <https://doi.org/10.1016/0025-5564(81)90043-2>
- Sanderson, M. J. 2002. Estimating absolute rates of molecular evolution and divergence times: a penalized likelihood approach. *Molecular Biology and Evolution*, 19(1), 101-109. <https://doi.org/10.1093/oxfordjournals.molbev.a003974>
- Smith, M. R. 2020. Information theoretic generalized Robinson–Foulds metrics for comparing phylogenetic trees. *Bioinformatics*, 36(20), 5007-5013. <https://doi.org/10.1093/bioinformatics/btaa614>
- Smith, S. A., & O'Meara, B. C. 2012. treePL: divergence time estimation using penalized likelihood for large phylogenies. *Bioinformatics*, 28(20), 2689-2690. <https://doi.org/10.1093/bioinformatics/bts492>
- Wright, E. S. 2024. Fast and Flexible Search for Homologous Biological Sequences with DECIPHER v3. *The R Journal*, 16(2), 191-200. <https://doi.org/10.32614/RJ-2024-022>
