---
title: "Tutorial 6: Core Pipeline Functions and Methodological Dictionary"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 6: Core Pipeline Functions and Methodological Dictionary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview and Methodological Dictionary

Following the step-by-step analytical workflows presented in Tutorials 1 through 5, this reference guide synthesizes the core pipeline functions and methodological principles implemented in `PhyloCactus`. While the complete technical documentation of all exported functions, parameters, and input/output signatures is provided in the standard R package manual pages (`pkgdown` reference section), this dictionary highlights the primary high-level modules that drive the sequential workflow, explaining their biological motivation and methodological rationale.

### Narrative Framework of the Reference Suite

- **Where are we?** Having assembled multilocus supermatrices, inferred maximum-likelihood phylogenies, estimated divergence times with `treePL`, and validated topologies against external backbones, we provide a unified dictionary detailing core pipeline functions and their evolutionary context.
- **Why are we here?** Complex plant phylogenetic pipelines require explicit documentation linking software parameters directly to underlying evolutionary mechanics, such as mitigating long-branch attraction (LBA), enforcing positional homology, and accounting for incomplete lineage sorting (ILS).
- **What will we achieve?** A structured methodological dictionary mapping sequence mining, alignment quality control, substitution model evaluation, maximum-likelihood search, temporal calibration, and comparative validation modules.
- **What comes next?** Users can query individual function manual pages (e.g., `?run_concatenation_pipeline`) or consult the `pkgdown` reference index for detailed parameter definitions and audit safeguards.

## Function Reference Dictionary

### Sequence Assembly & Mining

Functions in this category retrieve orthologous sequence clusters from public repositories using sequence similarity clustering (`phylotaR`), avoiding nomenclature ambiguities across public databases.

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `assemble_ingroup_phylotar()` | Mines orthologous DNA sequence clusters for a focal ingroup clade (e.g., Cactaceae; Guerrero et al. 2019) from GenBank using `phylotaR`. Applies strict taxonomic filters, evaluates cluster occupancy, and resolves nomenclature inconsistencies. | Bennett et al. (2018) |
| `assemble_outgroup_phylotar()` | Retrieves orthologous sequences for specified outgroup lineages to establish accurate root positions for downstream likelihood search and chronogram estimation. | Bennett et al. (2018) |
| `fetch_genbank_metadata()` | Interrogates NCBI Entrez utilities to extract sequence lengths, organism taxonomy, publication titles, and accession IDs for downloaded sequence clusters. | Camacho et al. (2009) |

### Alignment & Quality Control

Functions in this category establish hypotheses of positional homology (`MAFFT`) and apply automated masking of non-homologous or ambiguous regions (`DECIPHER`).

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `run_alignment_pipeline()` | Generates multiple sequence alignments using `MAFFT` and masks poorly aligned, ambiguous, or hypervariable regions using `DECIPHER`, eliminating systematic noise while retaining phylogenetically informative sites. | Katoh & Standley (2013); Wright (2024) |
| `run_marker_screening()` | Evaluates phylogenetic informativeness, sequence coverage, and locus quality across individual alignments, regressing uncorrected p-distance against a Gamma-corrected K80 distance. The slope threshold below which a locus is rejected is a `PhyloCactus` cleaning criterion, not a published saturation test; the citation covers the distance model only. | Kimura (1980); Jin & Nei (1990) |
| `clean_taxonomic_names()` | Standardizes tip labels against authoritative botanical checklists (e.g., Caryophyllales.org checklist), resolving synonymies and orthographic variants. | Korotkova et al. (2021) |
| `run_joint_realignment()` | Re-estimates positional homology alignments (`MAFFT`) after aggregating ingroup and outgroup sequence markers to refine alignment hypotheses. | Katoh & Standley (2013) |

### Concatenation & Supermatrix Assembly

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `run_concatenation_pipeline()` | Concatenates individual orthologous locus alignments into a unified multilocus supermatrix and generates partition coordinate maps compatible with `RAxML-NG`, `IQ-TREE`, and `MrBayes`. | Kozlov et al. (2019) |

### Likelihood Inference & Model Evaluation

Functions in this category select optimal nucleotide substitution models per partition and infer maximum-likelihood phylogenies under topological constraints.

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `preprocess_partitions()` | Validates structural integrity and syntax of the supermatrix and its corresponding partition definitions prior to likelihood evaluation using `RAxML-NG`. | Kozlov et al. (2019) |
| `run_modeltest_ng()` | Identifies the optimal substitution model per partition (`ModelTest-NG`) using AICc criteria, controlling for mutational rate heterogeneity and mitigating long-branch attraction (LBA). | Darriba et al. (2020) |
| `build_constraint_scaffold()` | Synthesizes a Newick multifurcating constraint scaffold enforcing monophyletic relationships of established higher taxonomic ranks. | Guerrero et al. (2019) |
| `calculate_ml_tree()` | Infers the maximum-likelihood evolutionary hypothesis explaining the supermatrix under selected partition models and constraint scaffolds (`RAxML-NG`). | Kozlov et al. (2019) |
| `map_branch_supports()` | Maps Felsenstein's Bootstrap Proportions (FBP) and Transfer Bootstrap Expectation (TBE) support values onto the best maximum-likelihood tree. | Felsenstein (1985); Lemoine et al. (2018) |

### Divergence Time Estimation & treePL Automation

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `rescale_tree()` | Multiplies tree edge lengths by a scaling factor prior to penalized likelihood dating (`treePL`) to improve numerical conditioning of rate parameter optimization. | Sanderson (2002) |
| `run_treePL_direct()` | Executes `treePL` natively via system calls to estimate ultrametric chronograms under penalized likelihood. | Smith & O'Meara (2012) |
| `run_treePL_cv()` | Executes cross-validation parameter optimization, rate smoothing selection, and dating stages from R. | Maurin (2020) |
| `automate_treePL()` | Automates cross-validation parameter optimization, rate smoothing selection, and chronogram estimation across temporal bootstrap replicates. | Maurin (2020) |
| `send_run_notification()` | Sends an email when a long local analysis ends, reporting status, timings and output paths. Never fails the run it reports on. | `blastula` R package, <https://github.com/rstudio/blastula> |

### Comparative Validation & IUCN Utilities

| Function | Biological & Methodological Application | Canonical Citation |
|:---|:---|:---|
| `validate_phylogenies()` | Quantifies topological congruence between the focal supermatrix tree and external published backbones using Robinson-Foulds distances and multidimensional scaling (MDS). | Robinson & Foulds (1981); Smith (2020) |
| `integrate_publication_tree()` | Maps statistical support onto chronograms, collapsing weakly supported nodes (FBP < 0.70) into soft polytomies to represent analytical uncertainty. | Hillis & Bull (1993) |
| `get_iucn_data()` | Interrogates the IUCN Red List API via `rredlist` to retrieve extinction threat categories, population trends, and geographic endemism metrics. | IUCN (2024) |
| `scale_fill_iucn()` | Provides a standardized `ggplot2` manual fill scale mapping official IUCN threat categories onto comparative figures. | IUCN (2024) |

### Reproducible Execution Example

```{r, eval = FALSE}
library(PhyloCactus)

# Query help documentation for any pipeline function
?run_alignment_pipeline
?calculate_ml_tree
?automate_treePL
```

## Conclusion

This dictionary covers the functions that drive the sequential workflow. The audit and verification functions that the tutorials call at each gate, including `check_calibration_consistency()`, `report_bound_adherence()`, `classify_constrained_nodes()`, `audit_calibration_support()` and `report_smoothing_sensitivity()`, are documented in their own manual pages and in the tutorials where they are used. Consult the `pkgdown` reference index for the complete list of exported functions and their signatures.

## References

- Bennett, D. J., Hettling, H., Silvestro, D., Zizka, A., Bacon, C. D., Faurby, S., ... & Antonelli, A. 2018. phylotaR: An automated pipeline for retrieving orthologous DNA sequences from GenBank in R. *Life*, 8(2), 20. <https://doi.org/10.3390/life8020020>
- Camacho, C., Coulouris, G., Avagyan, V., Ma, N., Papadopoulos, J., Bealer, K., & Madden, T. L. 2009. BLAST+: architecture and applications. *BMC Bioinformatics*, 10, 421. <https://doi.org/10.1186/1471-2105-10-421>
- Darriba, D., Posada, D., Kozlov, A. M., Stamatakis, A., Morel, B., & Flouri, T. 2020. ModelTest-NG: a new and scalable tool for the selection of DNA and protein evolutionary models. *Molecular Biology and Evolution*, 37(1), 291-294. <https://doi.org/10.1093/molbev/msz189>
- Felsenstein, J. 1985. Confidence limits on phylogenies: an approach using the bootstrap. *Evolution*, 39(4), 783-791. <https://doi.org/10.1111/j.1558-5646.1985.tb00420.x>
- Guerrero, P. C., Majure, L. C., Cornejo-Romero, A., & Hernández-Hernández, T. 2019. Phylogenetic Relationships and Evolutionary Trends in the Cactus Family. *Journal of Heredity*, 110(1), 4-21. <https://doi.org/10.1093/jhered/esy064>
- Hillis, D. M., & Bull, J. J. 1993. An empirical test of bootstrapping as a method for assessing confidence in phylogenetic analysis. *Systematic Biology*, 42(2), 182-192. <https://doi.org/10.1093/sysbio/42.2.182>
- IUCN. 2024. *The IUCN Red List of Threatened Species*. <https://www.iucnredlist.org>
- Jin, L., & Nei, M. 1990. Limitations of the evolutionary parsimony method of phylogenetic analysis. *Molecular Biology and Evolution*, 7(1), 82-102. <https://doi.org/10.1093/oxfordjournals.molbev.a040588>
- Katoh, K., & Standley, D. M. 2013. MAFFT multiple sequence alignment software version 7: Improvements in performance and usability. *Molecular Biology and Evolution*, 30(4), 772-780. <https://doi.org/10.1093/molbev/mst010>
- Kimura, M. 1980. A simple method for estimating evolutionary rates of base substitutions through comparative studies of nucleotide sequences. *Journal of Molecular Evolution*, 16(2), 111-120. <https://doi.org/10.1007/BF01731581>
- Korotkova, N., Aquino, D., Arias, S., Eggli, U., Franck, A., Gómez-Hinostrosa, C., Guerrero, P. C., Hernández, H. M., Kohlbecker, A., Köhler, M., Luther, K., Majure, L. C., Müller, A., Metzing, D., Nyffeler, R., Sánchez, D., Schlumpberger, B. O., & Berendsohn, W. G. 2021. Cactaceae at Caryophyllales.org, a dynamic online species-level taxonomic backbone for the family. *Willdenowia*, 51(2), 251-270. <https://doi.org/10.3372/wi.51.51208>
- Kozlov, A. M., Darriba, D., Flouri, T., Morel, B., & Stamatakis, A. 2019. RAxML-NG: a fast, scalable and user-friendly tool for maximum likelihood phylogenetic inference. *Bioinformatics*, 35(21), 4453-4455. <https://doi.org/10.1093/bioinformatics/btz305>
- Lemoine, F., Domelevo Entfellner, J. B., Wilkinson, E., Correia, D., Dávila Felipe, M., De Oliveira, T., & Gascuel, O. 2018. Renewing Felsenstein's phylogenetic bootstrap in the era of big data. *Nature*, 556(7702), 452-456. <https://doi.org/10.1038/s41586-018-0043-0>
- Maurin, K. J. L. 2020. An empirical guide for producing a dated phylogeny with treePL in a maximum likelihood framework. *arXiv*:2008.07054. <https://doi.org/10.48550/arXiv.2008.07054>
- Robinson, D. F., & Foulds, L. R. 1981. Comparison of phylogenetic trees. *Mathematical Biosciences*, 53(1-2), 131-147. <https://doi.org/10.1016/0025-5564(81)90043-2>
- Sanderson, M. J. 2002. Estimating absolute rates of molecular evolution and divergence times: a penalized likelihood approach. *Molecular Biology and Evolution*, 19(1), 101-109. <https://doi.org/10.1093/oxfordjournals.molbev.a003974>
- Smith, M. R. 2020. Information theoretic measures of phylogenetic tree distance. *Systematic Biology*, 69(6), 1168-1182. <https://doi.org/10.1093/sysbio/syaa016>
- Smith, S. A., & O'Meara, B. C. 2012. treePL: divergence time estimation using penalized likelihood for large phylogenies. *Bioinformatics*, 28(20), 2689-2690. <https://doi.org/10.1093/bioinformatics/bts492>
- Wright, E. S. 2024. Fast and Flexible Search for Homologous Biological Sequences with DECIPHER v3. *The R Journal*, 16(2), 191-200. <https://doi.org/10.18129/B9.bioc.DECIPHER>
