---
title: "Tutorial 2: Phylogenetic Inference and Divergence Time Estimation"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Phylogenetic Inference and Divergence Time Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 8,
  fig.align = "center"
)
```

## Abstract

Phylogenetic inference represents the transition from a curated molecular dataset to an explicit evolutionary hypothesis. After assembling and preparing multilocus sequence matrices in [Tutorial 1](https://beeamerino.github.io/PhyloCactus/articles/tutorial-1-cactus-phylogeny-prep.html), the next challenge is to estimate evolutionary relationships while accounting for the heterogeneous processes that shape molecular evolution.

This tutorial describes the second stage of the `PhyloCactus` workflow, where concatenated multilocus datasets are transformed into statistically evaluated phylogenetic hypotheses. The pipeline integrates partition specific substitution model selection, constrained maximum likelihood inference, assessment of topological uncertainty, and divergence time estimation using penalized likelihood.

Large multilocus phylogenetic datasets often contain substantial uncertainty due to incomplete molecular sampling, uneven taxonomic representation, and heterogeneous sequence availability. However, relationships among major taxonomic lineages may already be supported by previous phylogenetic studies. To incorporate this available biological knowledge while allowing the placement of newly sampled species, `PhyloCactus` uses a taxonomic constraint scaffold defined in `cactus_constraints.csv`. This file assigns sampled species to established higher level clades and subfamilies, providing a biologically informed framework that reduces exploration of unsupported regions of tree space while preserving inference among unresolved relationships within major lineages.

Divergence time estimation requires the integration of molecular branch lengths with temporal information derived from fossil evidence or previously estimated evolutionary frameworks. Because **Cactaceae** lacks reliable direct fossil calibrations suitable for dating the entire family, `PhyloCactus` implements temporal constraints through the `calibrations_bounds.csv` file. These bounds incorporate lineage specific temporal estimates from Hernández-Hernández *et al*. (2014) for major cactus clades and subfamilies, together with secondary calibration information for **Cactaceae** based on the broader flowering plant framework proposed by Ramírez-Barahona *et al*. (2020). This strategy allows the estimation of a temporal framework while acknowledging the limitations imposed by the fossil record.

Because evolutionary history cannot be directly observed, phylogenetic reconstruction relies on statistical models that approximate the processes generating molecular variation. Therefore, the resulting phylogeny should be interpreted as the evolutionary hypothesis that best explains the observed molecular data under the selected models and assumptions, rather than as an absolute representation of historical relationships.

The workflow combines `ModelTest-NG`, `RAxML-NG`, and `treePL` to estimate a time calibrated phylogeny from the curated supermatrices generated in the previous tutorial. The resulting evolutionary framework provides the foundation for downstream comparative analyses, including historical biogeography, diversification dynamics, and trait evolution.

## Complete Pipeline Execution Workflow

### Setup: Creating a Clean Workspace

Before proceeding with **Module 7**, ensure that the required external phylogenetic binaries (`ModelTest-NG`, `RAxML-NG`, and `treePL`) are installed on your machine and accessible via your system's `$PATH`.

*If you are using macOS, many of these can be installed via Homebrew (`brew install raxml-ng`, etc.). On Linux/Windows, ensure they are compiled and mapped correctly to your `$PATH`. Alternatively, if you receive a "command not found" error, you can provide the absolute path to the executable file (e.g., `raxml_path = "/usr/local/bin/raxml-ng"`) directly in the function arguments.*

To quickly run the analyses in this module, you can copy the entire executable script to your tutorial directory:

``` r
tutorial_dir <- "~/Desktop/PhyloCactus_Tutorial"
setwd(tutorial_dir)

# Copy the entire tutorial script for easy execution
file.copy(
  system.file("scripts", "tutorial-2-cactus-phylogeny-inference.R", package = "PhyloCactus"),
  file.path(tutorial_dir, "tutorial-2-cactus-phylogeny-inference.R")
)
```

### Module 7: Preprocess and Substitution Models

Phylogenetic datasets derived from multiple molecular markers often contain substantial evolutionary heterogeneity. Different loci may evolve under distinct substitution patterns, nucleotide compositions, and selective constraints. If these differences are ignored, a single evolutionary model applied across the entire dataset may inadequately describe the observed sequence variation and introduce biases during likelihood based phylogenetic inference.

Therefore, before estimating the phylogenetic relationships, the concatenated supermatrix must be represented within a statistical framework that accounts for variation among molecular partitions. Partition specific substitution models provide a more realistic approximation of sequence evolution by allowing each locus region to be analyzed according to its own evolutionary characteristics.

The **`preprocess_partitions()`** function prepares the concatenated supermatrix generated in Module 6 for downstream phylogenetic inference. It reads the alignment and partition files, verifies their structural compatibility using the `RAxML-NG` parser, and generates standardized input files required for substitution model evaluation and maximum likelihood analysis.

The stages of this tutorial run for hours and a complete analysis spans several sessions, so each one must be executable on its own rather than only as part of a single pass from top to bottom. A script that passes results between stages in variables cannot do that: restarting R and resuming at Module 8 clears `analysed_phy` and the stage fails on a missing object rather than on a missing file. Hard-coding the filenames instead removes the dependency on session state but reintroduces the problem the naming convention exists to solve, since the same literal name is then retyped at a dozen call sites and drifts out of step with the run prefix.

The **`resolve_run_paths()`** function reconstructs the absolute path of every file of the run from the output directory, the run prefix and the supermatrix location. Each module below opens by calling it, so every filename has a single definition while every stage remains independently runnable. Two of its entries are resolved by inspecting the run rather than by convention alone, because they record what happened rather than what was chosen: `analysed_phy` points at the reduced matrix when `RAxML-NG` collapsed identical terminals during validation and at the supermatrix when it did not, and `best_tree` points into `ml_search/` when the maximum likelihood search was submitted to a cluster and into the run root when it was executed locally. Passing `require` names the entries a stage depends on, turning a missing input into an error that identifies the module responsible for producing it.

The **`run_modeltest_ng()`** function performs automated substitution model evaluation for each predefined molecular partition. By interfacing with `ModelTest-NG` (Darriba *et al*., 2020; Flouri *et al*., 2015), the function compares alternative models of nucleotide evolution and identifies the model that provides the best statistical explanation of the observed sequence variation according to information criteria such as AICc (Akaike, 1974; Hurvich & Tsai, 1989). Selecting optimal partition-specific substitution models is crucial for controlling mutational rate heterogeneity across loci and mitigating systematic artifacts such as Long-Branch Attraction (LBA), where rapidly evolving non-homologous lineages false-group together. The resulting model scheme provides the statistical foundation for maximum-likelihood tree inference.

```{r, eval = FALSE}
library(PhyloCactus)

# --- SETUP: run this block first, in every session -----------------------------------------
output_dir <- "7_Phylogenetics"
dir.create(output_dir, showWarnings = FALSE)

# Get model test path. Note: Configure these in your .Renviron file
modeltest_path <- Sys.getenv("PATH_MODELTEST_NG", "modeltest-ng")
raxml_path     <- Sys.getenv("PATH_RAXML_NG", "raxml-ng")
treepl_path    <- Sys.getenv("PATH_TREEPL", "treePL")

# Every file of the run carries this prefix, so the run is renamed by changing one value.
run_prefix <- "cactus"

# Where Module 6 left the concatenated supermatrix.
supermatrix_file <- "6_Concatenated/concatenated_alignments/ALIGNMENT_supermatrix.phy"
partition_file   <- "6_Concatenated/concatenated_alignments/PARTITION_raxml_style.txt"

# Rooting terminals, declared once and reused by every step that needs them. The whole Talinaceae
# sample is used, not one terminal of it; see the note below for why. Taxon names are read from the
# first field of each PHYLIP record. Adapt `pattern` to the outgroup lineage sampled in your dataset.
supermatrix_taxa <- sub("\\s.*$", "", readLines(supermatrix_file)[-1])
rooting_outgroup <- resolve_rooting_outgroup(supermatrix_taxa[nzchar(supermatrix_taxa)],
                                             pattern = "^(Talinum|Talinella)_")
cat("Rooting on", length(rooting_outgroup), "terminals:",
    paste(rooting_outgroup, collapse = ", "), "\n")
# --- end of SETUP --------------------------------------------------------------------------

# model_handling = "force_dna" writes the datatype token DNA into the model field. RAxML-NG
# expands a bare DNA token into its own default (GTR+FC+G4m+B) and writes that expansion into the
# reduced partition file; handing that to ModelTest-NG would both pre-empt the model selection and
# crash its partition parser.
preprocess_partitions(
  phy_matrix = supermatrix_file,
  part_file = partition_file,
  raxml_path = raxml_path,
  output_dir = output_dir,
  prefix = run_prefix,
  force_check = FALSE,
  model_handling = "force_dna"
)

# Resolve every path of the run from the naming convention. RAxML-NG --check collapses terminals
# whose concatenated sequences are identical and writes a reduced matrix; those terminals are
# absent from the tree, so `paths$analysed_phy` points at the reduced matrix when one was written
# and at the supermatrix when it was not.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file)
print(paths)

dup_groups_file <- "6_Concatenated/logs_and_qc/SUPP_TABLE_identical_sequence_groups.csv"
if (file.exists(dup_groups_file)) {
  dup_groups <- utils::read.csv(dup_groups_file)
  cat("Identical-sequence groups exported by Module 6:", length(unique(dup_groups$group)),
      "covering", nrow(dup_groups), "terminals;",
      nrow(dup_groups) - length(unique(dup_groups$group)), "will be collapsed.\n")
}
if (isTRUE(paths$was_reduced)) {
  cat("Reduced matrix header (taxa sites):", readLines(paths$analysed_phy, n = 1), "\n")
} else {
  cat("No reduced matrix written: RAxML-NG kept every terminal of the supermatrix.\n")
}

# Run `ModelTest-NG` to select substitution models for each partition.
# ModelTest-NG resolves its output prefix relative to the working directory, so the call is made
# from inside output_dir. resolve_run_paths() already returns absolute paths, which matters because
# when RAxML-NG applied no reduction the analysed matrix still lives in 6_Concatenated.
setwd(output_dir)
run_modeltest_ng(
  modeltest_exec_path = modeltest_path,
  aln_file = paths$analysed_phy,
  part_file = paths$validated_part,
  prefix = paste0(run_prefix, "_modeltest"),
  threads = 8
)
setwd("..")
```

### Module 8: Constraint Trees and Maximum Likelihood Search

Maximum likelihood inference estimates the evolutionary hypothesis that provides the best explanation of the observed molecular data under a defined substitution model. The resulting phylogeny should not be interpreted as an absolute representation of biological history, but rather as the topology that maximizes the likelihood of observing the molecular alignment given the evolutionary models, constraints, and available molecular evidence.

Large phylogenetic datasets frequently include incomplete taxonomic sampling, uneven sequence availability, and uncertainty associated with the placement of newly incorporated species. These challenges become particularly relevant when expanding datasets to include a large number of taxa from highly diverse evolutionary radiations. In such cases, previously established phylogenetic knowledge can be incorporated to reduce uncertainty and improve the efficiency of tree space exploration.

The **`build_constraint_scaffold()`** function generates a topological constraint scaffold from the taxonomic framework provided in `cactus_constraints.csv`. This file assigns sampled species to major evolutionary groups within **Cactaceae**, including recognized subfamilies and internal clades. These assignments are based on previously published phylogenetic evidence and current evolutionary consensus regarding relationships among major cactus lineages, particularly the framework established by Guerrero *et al*. (2019).

The constraint scaffold is not intended to replace molecular inference or impose a complete predefined phylogeny. Instead, it incorporates well supported relationships among higher level lineages while allowing unresolved relationships, particularly among newly incorporated species, to be inferred directly from the molecular dataset. By constraining only relationships supported by previous evidence, the approach reduces uncertainty during maximum likelihood inference and focuses the analysis on the evolutionary relationships that remain to be resolved.

The **`calculate_ml_tree()`** function interfaces with `RAxML-NG` (Kozlov *et al*., 2019) to estimate the maximum likelihood topology from the concatenated molecular matrix using the partition specific substitution models identified in **Module 7**. The analysis searches for the tree configuration that maximizes the likelihood of observing the sequence data under the selected evolutionary framework while incorporating the constraint scaffold defined for **Cactaceae**.

The resulting maximum likelihood tree represents the best supported evolutionary hypothesis under the available molecular evidence, substitution models, and prior phylogenetic framework. This topology serves as the reference tree for subsequent branch support estimation and divergence time analyses.

The search runs either on a local workstation through `calculate_ml_tree()` or on an HPC allocation through **`generate_ml_search_script()`**, which writes a SLURM batch script performing the identical constrained search. The choice is made with the `run_local_ml` flag below, mirroring the `run_local_bs` switch used for bootstrap replicates in Module 9. Unlike the bootstrap step, the maximum likelihood search is submitted as a single job rather than a job array: `RAxML-NG` distributes the independent starting tree searches across workers within one job and writes a single best tree, so no collection step is required to compare log-likelihoods across tasks.

The parameter that governs wall-clock in this stage is `n_workers`, not the thread count. `RAxML-NG` assigns each worker `threads / workers` threads and gives it one starting tree at a time, so the search completes in `ceiling(n_init_trees / workers)` sequential rounds. Total core-work is fixed: adding threads to a single worker shortens each individual tree, while adding workers reduces the number of rounds, and the second lever dominates once each worker already holds an efficient thread count. A search of 50 starting trees on one worker executes 50 rounds; the same 50 trees across 25 workers execute two. In a reference single-worker benchmark run on an Apple M2 Pro (8 threads, `n_workers = 1`), 50 starting trees required 47115 s, approximately 938 s per tree, essentially all of it attributable to sequential rounds rather than to per-tree cost. In contrast, running across 25 parallel workers on an AMD EPYC 9754 server (75 threads, 3 threads per worker) on the full production supermatrix (1023 taxa, 12806 sites, 5954 patterns, 11 partitions) completed 50 starting trees in two rounds in only 2393 s (~40 minutes).

Passing `n_workers = NULL` derives the worker count automatically, constrained to a divisor of the starting tree count so that no worker sits idle in the final round. A worker count that does not divide the tree count wastes the tail of the search: 16 workers over 50 trees run four rounds with only two workers occupied in the last one, reaching the same wall-clock as 25 workers over two full rounds while holding more cores to do it. Per-worker thread allocation should keep several hundred site patterns per thread, since parallel efficiency within a worker saturates below that; `min_threads_per_worker` sets that floor and warrants re-validation on unfamiliar node hardware with a short trial search (`n_init_trees = "rand{2}"`) before committing a full allocation.

The **`calculate_rf_distances()`** function evaluates topological variability among independently inferred trees by calculating pairwise Robinson Foulds distances (Robinson & Foulds, 1981). These comparisons provide an additional measure of consistency among alternative tree searches and allow the assessment of whether independent analyses converge toward similar evolutionary hypotheses.

Rooting is imposed after the search, not during it, and it cannot be delegated to `RAxML-NG`. Under time-reversible substitution models the likelihood of a topology is identical for every one of its possible rootings (Felsenstein, 1981), so the data carry no information about where the root sits and no constraint tree can supply it: constraint trees, like the search itself, are unrooted. `--outgroup` only places the named terminals first in the output. The root is an outgroup decision, applied afterwards by `automate_treePL()` through `root_on_clade()`. What that argument does control is which edge the root sits on, and the choice governs which groups come out monophyletic, which nodes exist, and where each calibration lands, since `treePL` addresses nodes by the MRCA of the terminals declared for them.

Root on the whole outgroup clade rather than on one of its terminals. Naming a single terminal places the root *inside* the clade: the remaining terminals of that lineage fall on the ingroup side, the lineage is left paraphyletic in the final tree, and its crown node collapses onto the root. Any calibration addressed by the MRCA of that lineage then lands on the root instead of on the node it was written for. `resolve_rooting_outgroup()` returns the full set and `root_on_clade()` places the root on its stem edge, erroring when the terminals are not a clade of the unrooted topology. Use `root_on_clade()` rather than `ape::root()` directly: `RAxML-NG` writes the topology as an unrooted trifurcation, and when the rooting terminals fall on more than one of its basal branches, as they do in the reference dataset, `ape::root()` does not recognise them as a clade and leaves the root elsewhere.

For the reference Cactaceae dataset, the rooting set comprises the 8 sampled **Talinaceae** terminals (*Talinum* and *Talinella*). Occupancy across that set is reported per terminal in `6_Concatenated/final_tables/TABLE_final_species_alignment_summary.csv`. Rooting on Talinaceae places the root on the stem of the ACP clade (Anacampserotaceae, Cactaceae, Portulacaceae), avoiding the topological artifact of forcing an unverified sister relationship between Cactaceae and Anacampserotaceae that occurs when rooting on Portulacaceae in the absence of an external fourth lineage.

With the root placed on the stem of Talinaceae, the crown of the ACP clade is an internal node with an estimable stem branch length, allowing maximum-likelihood inference to test alternative topological resolutions among the three core families (Ramírez-Barahona *et al.*, 2020; Zuntini *et al.*, 2024; de Vos *et al.*, 2025). Interfamilial relationships within this rapid radiation represent a zone of topological conflict, and the recovered tree reflects empirical preference under the multilocus supermatrix rather than an assumption imposed by outgroup choice.

```{r, eval = FALSE}
# Module 8 needs the outputs of Module 7. Asking for them by name turns a missing input into an
# error that says which module has not run yet, instead of a missing object after a restart.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file,
                           require = c("analysed_phy", "best_models"))

# Assembles taxonomic classifications into a constraint scaffold
build_constraint_scaffold(
  alignment_path = paths$analysed_phy,
  constraints_csv_path = system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus"),
  output_dir = output_dir,
  prefix = run_prefix
)

# Execute the constrained maximum-likelihood search with `RAxML-NG`, either on this machine or as a
# SLURM job on an HPC allocation. Set run_local_ml to TRUE to search now, or FALSE to write a batch
# script to submit on the cluster.
run_local_ml <- FALSE

# Starting trees, declared once so that the local run and the cluster script search the same space.
ml_start_trees <- "rand{25},pars{25}"

if (run_local_ml) {
  # n_workers = NULL derives the worker count from threads and the starting tree count. Leaving it
  # at 1 puts every starting tree in its own sequential round, which is the single largest avoidable
  # cost in this stage.
  calculate_ml_tree(
    raxml_bin_path = raxml_path,
    aln_file = paths$analysed_phy,
    part_file = paths$best_models,
    constraint_file = paths$constraint_tree,
    outgroup = rooting_outgroup,
    n_init_trees = ml_start_trees,
    seed = 1111,
    n_workers = NULL,
    min_threads_per_worker = 4L,
    threads = 8,
    output_dir = output_dir,
    prefix = paste0(run_prefix, "_search")
  )
} else {
  ml_script <- generate_ml_search_script(
    alignment_file = paths$analysed_phy,
    partition_file = paths$best_models,
    constraint_file = paths$constraint_tree,
    outgroup = rooting_outgroup,
    n_init_trees = ml_start_trees,
    seed = 1111,
    # --- SLURM & HPC Resource Configuration ---
    cluster_job_name = "cactus_ml",
    cluster_partition = "main",                        # Partition name (e.g., "main", "standard", "general")
    cluster_nodes = 1L,                               # Number of compute nodes requested
    threads = 75,                                     # CPU cores requested (--cpus-per-task)
    cluster_mem = "16G",                              # RAM memory allocation (--mem, ~11 GB max used)
    cluster_time = "02:00:00",                        # Walltime limit (--time HH:MM:SS, ~41 min real runtime)
    cluster_mail_user = Sys.getenv("MY_EMAIL", ""),   # Notification email address (reads MY_EMAIL from .Renviron)
    load_module = c("gcc/14.2.0-nlhpc", "openmpi/5.0.3-o", "raxml-ng/1.1.0-mpi-zen4-n"), # Cluster environment modules
    raxml_exec = "raxml-ng-mpi",                      # Executable binary in $PATH
    # --- Search Tuning ---
    workers = NULL,
    min_threads_per_worker = 3L,
    preparse = TRUE,
    output_dir = output_dir,
    prefix = paste0(run_prefix, "_search")
  )
  cat("Submit on cluster with: cd ml_search && sbatch run_ml_search.sh\n")
  cat("When the job finishes, sync results back and resume from the Module 9 block below.\n")
}

# Re-resolve now that Module 8 has written its outputs. `resolve_run_paths()` finds the best tree in
# `ml_search/` when the search ran on the cluster and in the run root when it ran locally, so every
# step below is identical either way.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file)

# Calculate Robinson-Foulds distance among maximum-likelihood trees
cat("Calculating distances among maximum-likelihood trees...\n")
rf_dist <- calculate_rf_distances(
  raxml_bin_path = raxml_path,
  ml_trees_file = paths$ml_trees,
  output_dir = output_dir
)
```

### Module 9: Topological Validation via Bootstrap Support

Once the maximum likelihood topology has been inferred, the statistical support of recovered clades must be evaluated by quantifying topological stability associated with each internal node. Bootstrap analyses provide an empirical measure of topological stability by assessing how consistently specific relationships are recovered under resampling of the original sequence alignment.

The **`generate_bootstrap_script()`** and **`run_local_bootstraps()`** functions configure and execute bootstrap searches using `RAxML-NG`. These procedures generate independent resampled datasets while maintaining the same substitution model, constraint tree and partition scheme defined during the maximum likelihood inference. The resulting bootstrap trees represent alternative evolutionary hypotheses derived from the molecular dataset and allow the assessment of uncertainty across the inferred topology.

The **`collect_bootstraps()`** function gathers the independent bootstrap searches, while **`check_bs_convergence()`** evaluates whether a sufficient number of replicates has been generated using the bootstopping criterion (Pattengale *et al*., 2010). Convergence is evaluated under Felsenstein's Bootstrap Proportions (`FBP`), testing whether the autoMRE threshold (cutoff <= 0.03) is satisfied across resampled replicates.

The **`map_branch_supports()`** function summarizes bootstrap information and transfers support values onto the best maximum likelihood topology. `PhyloCactus` adopts Felsenstein's Bootstrap Proportions (`FBP`) as its primary canonical metric. Published phylogenetic studies in Cactaceae (Hernández-Hernández *et al*., 2014; Arakaki *et al*., 2011) report standard FBP bipartition frequencies or Bayesian posterior probabilities. Transfer Bootstrap Expectation (`TBE`; Lemoine *et al*., 2018) is computed as a secondary, complementary metric. While TBE mitigates the penalization caused by unstable wildcard taxa in large supermatrices with missing data, it systematically inflates support values in large clades compared to standard bipartition frequencies (for example, free clades of 51 to 200 tips exhibit a median FBP of 0.265 alongside a median TBE of 0.969). Reporting FBP as the primary metric ensures direct comparability with published botanical benchmarks, while retaining TBE provides diagnostic insight into local taxon placement stability.

When interpreting support values, internal nodes enforced by the taxonomic constraint scaffold (`cactus_constraints.tree`) return support values of 1.000 by algorithmic definition across all bootstrap replicates. These values represent the topological constraint itself rather than empirical evidence from the molecular supermatrix, and must be designated as constrained rather than estimated in publication tables.

#### Temporal Bootstrap Replicates for Divergence Time Estimation

Divergence time estimation requires an additional assessment of uncertainty because branch lengths inferred during maximum likelihood analysis are influenced by sampling variation in the molecular data. Following the empirical protocol proposed by Maurin (2020) for dating large phylogenies with `treePL`, `PhyloCactus` generates a second set of bootstrap replicates specifically designed for temporal analyses.

The **`calculate_temporal_bootstraps()`** function generates bootstrap alignments and performs maximum likelihood searches while constraining the topology to the previously inferred maximum likelihood tree. This constraint does not force branch lengths to be identical; instead, it preserves the supported evolutionary relationships while allowing each bootstrap replicate to independently estimate branch length variation from the resampled molecular data.

This strategy is particularly important for large phylogenetic datasets where many relationships are already strongly supported at higher taxonomic levels, but uncertainty remains associated with branch length estimation and the placement of newly incorporated taxa. By maintaining a common topological framework, temporal bootstrap replicates provide a distribution of alternative branch length estimates that can be propagated into downstream divergence time analyses.

The resulting bootstrap trees are subsequently used by the dating module with `treePL` (Smith & O’Meara, 2012). Following the protocol described by Maurin (2020), the best maximum likelihood topology is first optimized to determine the appropriate penalized likelihood parameters and smoothing value. These parameters are then applied to the temporal bootstrap replicates to estimate a distribution of dated trees, from which confidence intervals for node ages can be obtained.

Therefore, `PhyloCactus` distinguishes between two complementary bootstrap procedures: standard bootstrap analyses evaluate the topological stability of inferred relationships, whereas temporal bootstrap replicates quantify uncertainty in molecular branch length estimates required for divergence time estimation.

```{r, eval = FALSE}
# -------------------------------------------------------------
# Estimate Bootstrap Replicates
# -------------------------------------------------------------
# Resuming here after a restart: run the SETUP block of Module 7, then continue from this line.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file,
                           require = c("analysed_phy", "best_models",
                                       "constraint_tree"))

# `RAxML-NG` supports running Bootstraps either locally or via HPC array chunks.
# You can set run_local_bs to TRUE to run them now on your machine, 
# or FALSE to generate an HPC bash script to run them remotely.
run_local_bs <- FALSE

if (run_local_bs) {
  cat("\n1. Estimating bootstraps locally...\n")
  bs_dir <- file.path(output_dir, "local_bs")
  local_bs <- run_local_bootstraps(
    raxml_bin_path = raxml_path,
    aln_file = paths$analysed_phy,
    part_file = paths$best_models,
    constraint_file = paths$constraint_tree,
    bs_trees = 500,
    outgroup = rooting_outgroup,
    threads = 8,
    workers = 1,
    output_dir = bs_dir
  )
} else {
  cat("\n1. Generating BS script for HPC chunks...\n")
  bs_dir <- file.path(output_dir, "bs_chunks")
  bs_script <- generate_bootstrap_script(
    alignment_file = paths$analysed_phy,
    partition_file = paths$best_models,
    constraint_file = paths$constraint_tree,
    outgroup = rooting_outgroup,
    # --- SLURM & HPC Resource Configuration ---
    cluster_job_name = "cactus_bs",
    cluster_partition = "main",                        # Partition name (e.g., "main", "standard", "general")
    cluster_nodes = 1L,                               # Number of compute nodes requested
    threads = 40,                                     # CPU cores per array task (--cpus-per-task)
    cluster_mem = "10G",                              # RAM memory allocation per task (--mem, ~5.3 GB max used)
    cluster_time = "24:00:00",                        # Walltime limit per task (--time HH:MM:SS, ~17 h real runtime)
    cluster_mail_user = Sys.getenv("MY_EMAIL", ""),   # Notification email address (reads MY_EMAIL from .Renviron)
    load_module = c("gcc/14.2.0-nlhpc", "openmpi/5.0.3-o", "raxml-ng/1.1.0-mpi-zen4-n"), # Cluster environment modules
    raxml_exec = "raxml-ng-mpi",                      # Executable binary in $PATH
    # --- Bootstrap Array Settings ---
    bs_per_rep = 500,                                 # Bootstrap replicates per array task
    max_reps = 2,                                     # Number of array tasks (2 x 500 = 1000 total replicates)
    workers = 10,
    output_dir = file.path(output_dir)
  )
  cat("Submit on cluster with: cd bs_chunks && sbatch run_bs_chunks.sh\n")
  cat("When all array tasks finish, sync results back and continue with collect_bootstraps().\n")
}

# -------------------------------------------------------------
# Collect Bootstrap Replicates
# -------------------------------------------------------------
# Before running this step, ensure that your bootstrap replicates have finished.
# If you ran them on HPC, make sure all chunks completed successfully and
# optionally sync the files back to your local repository.

cat("\n2. Collecting bootstrap replicates...\n")
# bs_dir was defined in Step 1 depending on whether you ran locally or via HPC chunks
bs_concat_file <- collect_bootstraps(
  bs_dir = bs_dir,
  output_dir = output_dir,
  prefix = paste0(run_prefix, "_ALL_bootstraps")
)
cat("   Bootstraps concatenated to:", bs_concat_file, "\n")

# Pick up the file collect_bootstraps() just wrote.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file,
                           require = c("all_bootstraps", "best_tree"))

# -------------------------------------------------------------
# Check Bootstrap Convergence
# -------------------------------------------------------------
# autoMRE bootstopping criterion evaluated under FBP (Felsenstein's Bootstrap Proportions)
cat("\n3. Checking bootstrap convergence (FBP autoMRE)...\n")
converge_log <- check_bs_convergence(
  raxml_bin_path = raxml_path,
  bs_trees_file = paths$all_bootstraps,
  bs_metric = "fbp",
  bs_cutoff = 0.03,
  seed = 111,
  threads = 8,
  output_dir = output_dir,
  prefix = paste0(run_prefix, "_bs_convergence")
)
print(converge_log)

# -------------------------------------------------------------
# Map Bootstrap Supports onto Best Tree (FBP Primary, TBE Secondary)
# -------------------------------------------------------------
cat("\nMapping bootstrap supports onto the best tree...\n")

# Primary metric: FBP support (comparable with published literature)
support_tree <- map_branch_supports(
  raxml_bin = raxml_path,
  best_tree = paths$best_tree,
  bootstraps_file = paths$all_bootstraps,
  metric = "fbp",
  output_dir = output_dir,
  prefix = paste0(run_prefix, "_support")
)
cat("   Primary FBP support tree generated at:", support_tree, "\n")

# Secondary metric: TBE support (complementary missing-data assessment)
support_tree_tbe <- map_branch_supports(
  raxml_bin = raxml_path,
  best_tree = paths$best_tree,
  bootstraps_file = paths$all_bootstraps,
  metric = "tbe",
  output_dir = output_dir,
  prefix = paste0(run_prefix, "_support_tbe")
)
cat("   Secondary TBE support tree generated at:", support_tree_tbe, "\n")

# Note: for Dating, run sequential temporal bootstraps constrained over the best tree.
run_local_temporal_bs <- FALSE  # Set to TRUE to execute locally on your machine

if (run_local_temporal_bs) {
  cat("Running temporal bootstraps locally on machine...\n")
  temporal_bs <- calculate_temporal_bootstraps(
    raxml_bin_path = raxml_path,
    aln_file = paths$analysed_phy,
    part_file = paths$best_models,
    best_tree_file = paths$best_tree,
    outgroup = rooting_outgroup,
    bs_trees = 100, # Adjust to 500 or 1000 for production.
    threads = 8,
    workers = 2,
    output_dir = paths$temporal_bs_dir,
    prefix = paths$temporal_bs_prefix
  )
} else {
  cat("Generating HPC SLURM submission script for temporal bootstraps...\n")
  hpc_temp_bs_script <- generate_temporal_bootstrap_script(
    alignment_file = paths$analysed_phy,
    partition_file = paths$best_models,
    best_tree_file = paths$best_tree,
    outgroup = rooting_outgroup,
    bs_trees = 500,
    threads = 40,
    workers = 10,
    threads_per_worker = 4L,
    preparse = TRUE,
    output_dir = output_dir,
    script_name = "run_temporal_bs.sh",
    cluster_job_name = "cactus_temp_bs",
    cluster_partition = "main",
    cluster_nodes = 1L,
    cluster_mem = "16G",
    cluster_time = "04:00:00",
    cluster_mail_user = Sys.getenv("MY_EMAIL", "")
  )
  cat("   Temporal bootstrap HPC script generated at:", hpc_temp_bs_script, "\n")
  cat("   Submit on cluster: cd cactus_temporal_bs && sbatch run_temporal_bs.sh\n")
}
```

### Module 10: Estimating Divergence Times via Penalized Likelihood

The inferred maximum likelihood topology provides a robust hypothesis of evolutionary relationships; however, branch lengths estimated from molecular data represent substitutions per site rather than geological time. To reconstruct the temporal history of **Cactaceae** diversification, the phylogenetic tree must therefore be transformed into an ultrametric chronogram by integrating molecular branch length information with temporal constraints derived from previous evolutionary studies.

`PhyloCactus` estimates divergence times using the penalized likelihood framework implemented in `treePL` (Sanderson, 2002; Smith & O’Meara, 2012). This approach is particularly suitable for large phylogenetic datasets because it accommodates lineage specific variation in evolutionary rates while remaining computationally feasible for analyses involving hundreds or thousands of taxa.

Temporal constraints are provided through the `calibrations_bounds.csv` reference file distributed with the package. Each row declares the node it constrains as a set of taxa (`column` and `value`), the kind of node it represents (`node_type`, either `stem` or `crown`), the minimum and maximum age bounds, and the study the bounds were taken from. `treePL` addresses nodes by the MRCA of the terminals listed for them, so the taxon set, not the row label, is what determines where a bound lands.

The distinction between stem and crown nodes governs which bound may be assigned to which node. A crown node is the first divergence within a clade; a stem node is the divergence separating that clade from its sister, and is therefore always older. Assigning a stem age to a crown node inflates the node by construction, and propagates that inflation to every node below it.

Internal calibrations for major cactus lineages are derived from Hernández-Hernández *et al.* (2014), which estimated divergence times among the principal evolutionary groups of the family. These are crown ages of densely sampled clades (**Opuntioideae**, **Cactoideae**, Cacteae) and are applied to the corresponding crown nodes.

Because **Cactaceae** lacks an unequivocal macrofossil record suitable for direct internal node calibration, the root of the ingroup is constrained by secondary calibration from the angiosperm-wide chronogram of Ramírez-Barahona *et al.* (2020) under the `RC_complete` scheme. With Talinaceae (*Talinum* and *Talinella*) explicitly sampled to root the tree, the calibrated node is:

- `ACP_root`: Constrains the crown divergence of the Anacampserotaceae + Portulacaceae + **Cactaceae** clade (corresponding to the stem node of Portulacaceae in Ramírez-Barahona *et al.*, 2020). The primary calibration scheme (Scheme S1) applies a fixed point of 41.82 Ma (the median stem age under BEAST relaxed clock), with alternative sensitivity schemes available in `calibrations_bounds.csv` (Scheme S2: 95% HPD interval [27.81, 53.37] Ma; Scheme S4: 53.22 Ma from Zuntini *et al.*, 2024 young tree).
- `Cactaceae_Anacampserotaceae_stem` ([21.95, 48.51] Ma) is retained in `calibrations_bounds.csv` with `used_in_analysis = FALSE` because it assumes a sister relationship between **Cactaceae** and Anacampserotaceae; our unconstrained likelihood inference and phylogenomic evidence (Zuntini *et al.*, 2024 Kew Tree of Life Release 4.0) instead recover (Anacampserotaceae + Portulacaceae) sister to **Cactaceae**.

The crown intervals reported by Ramírez-Barahona *et al.* (2020) for these families are also retained with `used_in_analysis = FALSE`, for provenance rather than for use. That study samples two terminals per family, so its crown estimate is the divergence between the two sampled species rather than the crown of the family, and systematically underestimates it. Its stem nodes are well defined regardless of within-family sampling, which is why the stem bounds are the ones carried forward.

Secondary calibration inherits the uncertainty of the study it comes from without propagating it, so the resulting node ages are conditional on that framework rather than independent estimates. Bounds drawn from Hernández-Hernández *et al.* (2014) and from Ramírez-Barahona *et al.* (2020) also come from analyses calibrated under different schemes. Both limitations should be stated in Methods.

The **`automate_treePL()`** function performs the complete divergence time estimation workflow. First, it prepares the maximum likelihood topology, branch length information, and calibration constraints required by `treePL`. The function then performs the optimization procedure required to estimate appropriate penalized likelihood parameters, including the smoothing parameter selected through cross validation.

Following the empirical protocol described by Maurin (2020), the optimization and cross validation steps are performed using the best scoring maximum likelihood tree without bootstrap annotations. This approach ensures that the dating parameters are optimized using the primary evolutionary hypothesis before being propagated to replicate datasets. Specifically, the smoothing value selected by cross validation on the maximum likelihood tree is reused, unmodified, for every temporal bootstrap replicate, rather than re-optimized independently for each one. Repeating full cross validation for every replicate is computationally prohibitive at typical bootstrap replicate counts, and reusing a single, well-supported smoothing value is standard practice in `treePL`-based dating pipelines; the trade-off is that per-replicate variation in optimal rate heterogeneity smoothing is not itself propagated into the reported confidence intervals, which instead capture uncertainty arising from branch length resampling alone.

After the optimal `treePL` parameters have been identified, the dating procedure is applied to the temporal bootstrap replicates generated in **Module 9**. These replicates were obtained by resampling the original alignment while maintaining the maximum likelihood topology as a constraint. This strategy preserves the inferred relationships among taxa while allowing branch lengths to vary according to molecular sampling uncertainty.

Propagating temporal uncertainty through these bootstrap replicates allows divergence time estimates to incorporate variation in branch length estimation. The resulting collection of dated bootstrap trees represents uncertainty associated with the temporal reconstruction under a maximum likelihood framework and provides the basis for estimating confidence intervals around divergence times.

#### Summarizing Dated Temporal Bootstrap Replicates

After all temporal bootstrap replicates have been independently dated, the resulting chronograms must be summarized into a single representative time calibrated tree. Following the empirical workflow proposed by Maurin (2020), `PhyloCactus` uses `TreeAnnotator` (Helfrich *et al.*, 2018) to summarize the dated bootstrap distribution.

Because temporal bootstrap replicates are generated under the maximum likelihood topology constraint, all replicate trees share identical branching relationships. Therefore, the summary procedure does not evaluate alternative topologies, but instead estimates the distribution of node ages across independently dated replicate trees.

The recommended `TreeAnnotator` configuration specifies a maximum sum of clade credibility target tree type, mean height node age estimates, and 0% burn-in.

Under this framework, `TreeAnnotator` calculates the mean age of each node and summarizes the uncertainty associated with divergence time estimation across the dated temporal bootstrap replicates.

TreeAnnotator reports these uncertainty intervals as highest posterior density (HPD) values due to compatibility with Bayesian phylogenetic workflows. However, these values should not be interpreted as Bayesian posterior distributions. Within the `PhyloCactus` framework, they represent confidence intervals derived from variation among independently dated temporal bootstrap replicates generated through penalized likelihood optimization.

Here is how you execute, print, and visualize these chronological calculations:

```{r, eval = FALSE}
# Resuming here after a restart: run the SETUP block of Module 7, then continue from this line.
paths <- resolve_run_paths(output_dir, run_prefix, supermatrix_file,
                           require = c("analysed_phy", "best_tree", "temporal_bs"))

dating_dir <- "8_Dating"
dir.create(dating_dir, showWarnings = FALSE)

# 1. Prepare Calibrations
cat("Creating `treePL` calibrations configurations...\n")
calibs_all <- read.csv(system.file("extdata", "calibrations_bounds.csv", package = "PhyloCactus"))
calibs <- calibs_all[calibs_all$used_in_analysis == TRUE, ]

constraints <- read.csv(system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus"))
ml_tree <- ape::read.tree(paths$best_tree)
tip_labels <- ml_tree$tip.label

# `numsites` must be the length of the matrix RAxML-NG actually analysed, not a rounded figure.
# treePL uses it to convert branch lengths into expected substitution counts, so it is read from
# the header of the matrix Module 7 reported as analysed. `paths$analysed_phy` already resolves to the
# reduced PHYLIP when terminals were collapsed and to the supermatrix when they were not, so no
# second guess at the filename is needed here.
phy_header <- strsplit(trimws(readLines(paths$analysed_phy, n = 1)), "\\s+")[[1]]
num_sites <- as.integer(phy_header[2])
cat("   Matrix analysed:", paths$analysed_phy, "-", phy_header[1], "taxa x", num_sites, "sites\n")

# Rooting terminals, declared once in Module 8 and re-derived here so this block also runs
# standalone after a restart.
if (!exists("rooting_outgroup")) rooting_outgroup <- resolve_rooting_outgroup(tip_labels, pattern = "^(Talinum|Talinella)_")

# `treePL` resolves every `mrca` entry by the MRCA of the listed terminals, so a correct
# label is no guarantee of a correct node. The rooted topology is reconstructed here to
# verify node identity before any bound is written.
rooted_ml <- root_on_clade(ml_tree, rooting_outgroup)
root_node <- ape::Ntip(rooted_ml) + 1L

# `value` accepts a semicolon-separated list, which is required by the two stem
# calibrations: a stem node is shared by the families it subtends and cannot be addressed
# by a single family name.
tips_for <- function(column, value) calibration_tips(constraints, column, value, tip_labels)

# Bounds are declared per node, but the nodes are nested and an ultrametric tree forces every
# ancestor to be older than each of its descendants. A set that is defensible row by row can still
# describe a tree that cannot exist, and treePL does not report the violation: it pins the
# offending node to a bound and returns a chronogram that reads like an estimate.
check_calibration_consistency(calibs, rooted_ml, constraints)

cfg_lines <- c()
for (i in seq_len(nrow(calibs))) {
  row <- calibs[i, ]
  tips_in_tree <- tips_for(row$column, row$value)

  if (length(tips_in_tree) < 2) next

  mrca_line <- paste("mrca =", row$mrca, paste(tips_in_tree, collapse = " "))
  min_line  <- sprintf("min = %s %f", row$mrca, row$min)
  max_line  <- sprintf("max = %s %f", row$mrca, row$max)

  cfg_lines <- c(cfg_lines, mrca_line, min_line, max_line)
}

# Node-identity assertions, rewritten when Talinaceae entered the sampling.
#
# With Talinaceae rooting the tree, the root is the crown of the ACPT clade and ACP_root is an
# internal node with a parent branch, which is what makes it identifiable rather than a parameter
# parked at the end of the tree. The earlier assertion required ACP_root to BE the root and would
# now fail, correctly: it encoded the previous sampling.
#
# What still has to hold, and what these lines check:
#   1. The rooting clade is monophyletic. Rooting on a single terminal instead of the clade leaves
#      the lineage paraphyletic and collapses its crown onto the root, which silently reassigns
#      every bound addressed by that lineage.
#   2. ACP_root resolves to a node that is NOT the root, so the anchor is placed on an internal
#      node rather than on the deepest split of the tree.
#   3. ACP_root is monophyletic. If the likelihood placed Talinaceae inside the ACP clade, the
#      anchor would be addressing something other than the node it was written for.
stopifnot(ape::is.monophyletic(rooted_ml, grep("^(Talinum|Talinella)_", tip_labels, value = TRUE)))
acp_tips <- tips_for("Family", "Cactaceae;Anacampserotaceae;Portulacaceae")
stopifnot(ape::getMRCA(rooted_ml, acp_tips) != root_node)
stopifnot(ape::is.monophyletic(rooted_ml, acp_tips))

# The resolution of the ACPT quartet is a result of this run, not an assumption, so it is reported
# rather than asserted. Whether Cactaceae_Anacampserotaceae_stem can be reactivated in
# calibrations_bounds.csv depends on what this prints.
cact_anac <- tips_for("Family", "Cactaceae;Anacampserotaceae")
anac_por  <- tips_for("Family", "Anacampserotaceae;Portulacaceae")
cact_por  <- tips_for("Family", "Cactaceae;Portulacaceae")
cat("ACPT quartet resolution recovered by this run:\n")
cat("   Cactaceae + Anacampserotaceae monophyletic: ",
    ape::is.monophyletic(rooted_ml, cact_anac), "\n")
cat("   Anacampserotaceae + Portulacaceae monophyletic: ",
    ape::is.monophyletic(rooted_ml, anac_por), "\n")
cat("   Cactaceae + Portulacaceae monophyletic: ",
    ape::is.monophyletic(rooted_ml, cact_por), "\n")

# `num_sites` here is the true length of the analysed matrix, which is what belongs in a file
# meant to document the analysis. `automate_treePL()` rescales branch lengths and divides this
# value by the same factor before handing a copy to `treePL`, so the substitution count per
# branch stays correct. The copy it actually runs is written to `auto_results/ML_tree/`.
treepl_cfg <- c(
  paste0("numsites = ", num_sites),
  cfg_lines,
  "nthreads = 8",
  "thorough"
)

calibrations_cfg_path <- file.path(dating_dir, "calibrations_treePL_fulltips.cfg")
writeLines(treepl_cfg, calibrations_cfg_path)
cat("   Calibrations compiled successfully to:", calibrations_cfg_path, "\n")

# 2. Run Fast Automated `treePL` Dating Pipeline
cat("\nRunning automated `treePL` wrapper script over maximum-likelihood tree and bootstrap replicates...\n")

# Priming, cross-validation and dating follow the empirical protocol of Maurin (2020,
# arXiv:2008.07054, CC BY 4.0) and run through run_treePL_cv(), in R. Until 2026-09-02 they were
# driven by a copy of the shell script at https://github.com/tongjial/treepl_wrapper, which was an
# earlier implementation of the same protocol; that script carries no licence, so it could not be
# redistributed inside this GPL-3 package, and it wrote `smoothing = ` where treePL reads
# `smooth = `, silently dating every chronogram at the built-in default of 10.
#
# Three defaults follow Maurin where the shell script did not: the priming parameters are the
# lowest rather than the most frequent, cross-validation uses `randomcv` rather than leave-one-out
# `cv`, and the smoothing grid reaches 1e-08 rather than stopping at 1e-04. Maurin reports optimal
# smoothing between 1e-06 and 1e-08 for a tree whose branch lengths were rescaled as they are here.

# Note: For tutorial purposes, you can limit the number of bootstrap trees to process
# by setting `num_bs = 100` (or any other number). If not provided, it will process all
# available bootstrap trees. Here we set it to 100 for faster tutorial execution.
automate_treePL(
  cfg_file = calibrations_cfg_path,
  ml_tree_file = paths$best_tree,
  bs_trees_file = paths$temporal_bs,
  numsites = num_sites,
  outgroup = rooting_outgroup,
  results_dir = file.path(dating_dir, "auto_results"),
  treePL_out = dating_dir,
  num_bs = 100
)

# 3. View and evaluate dating metrics cleanly in R
cat("\n--- Chronological Dating Results Summary ---\n")
if (file.exists(file.path(dating_dir, "BestTree_treePL.tree"))) {
  ml_chronogram <- ape::read.tree(file.path(dating_dir, "BestTree_treePL.tree"))
  cat("   Best ML Chronogram:\n")
  cat("     - File path:", file.path(dating_dir, "BestTree_treePL.tree"), "\n")
  cat("     - Root age:", max(ape::node.depth.edgelength(ml_chronogram)), "Mya\n")
  cat("     - Number of tips:", length(ml_chronogram$tip.label), "\n")

  # A node whose age equals one of its own bounds was not estimated: penalized likelihood returned
  # the constraint, and the chronogram gives no sign of it. ACP_root is fixed (min = max), so it
  # sits on its bound by construction; the other three calibrations are expected to be estimated.
  # The bootstrap chronograms are passed deliberately: a single point estimate can land just inside
  # a bound while the underlying age is unidentifiable, and only the replicates separate the two.
  bs_chronograms <- file.path(dating_dir, "bsTree_treePL.tree")
  adherence <- report_bound_adherence(
    ml_chronogram, calibs, constraints,
    bootstraps = if (file.exists(bs_chronograms)) bs_chronograms else NULL
  )
  cat("\n   Calibrated node ages against their bounds:\n")
  print(adherence)
}
cat("--------------------------------------------\n")
```

## Conclusion

At this stage, phylogenetic inference and divergence time estimation have been completed. The resulting maximum likelihood topology, branch support estimates, and time calibrated chronogram provide the evolutionary framework required for downstream comparative analyses.

The next stage of the `PhyloCactus` workflow focuses on integrating phylogenetic outputs with biological metadata and generating publication quality visualizations. In the following tutorial, users will learn how to visualize the maximum likelihood phylogeny with branch support values, display the dated chronogram with temporal uncertainty estimates, and integrate external biodiversity information, including IUCN Red List conservation categories.

[Continue to Tutorial 3: Data Visualization and IUCN Summaries](https://beeamerino.github.io/PhyloCactus/articles/tutorial-3-cactus-phylogeny-visualization.html)

## References

- Akaike, H. 1974. A new look at the statistical model identification. *IEEE Transactions on Automatic Control*, 19(6), 716-723. <https://doi.org/10.1109/TAC.1974.1100705>
- Arakaki *et al*. 2011. Contemporaneous and recent radiations of the world’s major succulent plant lineages. *Proceedings of the National Academy of Sciences of the United States of America*, 108(20), 8379–8384. <https://doi.org/10.1073/pnas.1100628108>
- Darriba *et al*. 2020. ModelTest-NG: a new and scalable tool for the selection of DNA and protein evolutionary models. *Molecular Biology and Evolution*, 37(1), 291-294. <https://doi.org/10.1093/molbev/msz189>
- Felsenstein, J. 1981. Evolutionary trees from DNA sequences: A maximum likelihood approach. *Journal of Molecular Evolution*, 17(6), 368–376. <https://doi.org/10.1007/BF01734359>
- Flouri *et al*. 2014. The Phylogenetic Likelihood Library. *Systematic Biology*, 64(2), 356-362. <https://doi.org/10.1093/sysbio/syu084>
- Helfrich *et al*. 2018. TreeAnnotator: versatile visual annotation of hierarchical text relations. *Proceedings of the Eleventh International Conference on Language Resources and Evaluation*. <https://lrec.elra.info/lrec2018-main-308>
- Hernández-Hernández *et al*. 2014. Beyond aridification: Multiple explanations for the elevated diversification of cacti in the New World Succulent Biome. *New Phytologist*, 202(4), 1382–1397. <https://doi.org/10.1111/nph.12752>
- Kozlov *et al*. 2019. RAxML-NG: A fast, scalable and user-friendly tool for maximum likelihood phylogenetic inference. *Bioinformatics*, 35(21), 4453–4455. <https://doi.org/10.1093/bioinformatics/btz305>
- Lemoine *et al*. 2018. Renewing Felsenstein’s phylogenetic bootstrap in the era of big data. *Nature*, 556(7702), 452-456. <https://doi.org/10.1038/s41586-018-0043-0>
- Maurin, K. J. 2020. An empirical guide for producing a dated phylogeny with treePL in a maximum likelihood framework. *arXiv preprint arXiv:2008.07054*. <https://doi.org/10.48550/arXiv.2008.07054>
- Pattengale *et al*. 2010. How many bootstrap replicates are necessary?. *Journal of Computational Biology*, 17(3), 337–354. <https://doi.org/10.1089/cmb.2009.0179>
- Ramírez-Barahona *et al*. 2020. The delayed and geographically heterogeneous diversification of flowering plant families. *Nature Ecology and Evolution*, 4(9), 1232–1238. <https://doi.org/10.1038/s41559-020-1241-3>
- Robinson, D. F., & Foulds, L. R. 1981. Comparison of phylogenetic trees. *Mathematical Biosciences*, 53(1-2), 131-147. <https://doi.org/10.1016/0025-5564(81)90043-2>
- Sanderson, M. J. 2002. Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach. *Molecular Biology and Evolution*, 19(1), 101–109. <https://doi.org/10.1093/oxfordjournals.molbev.a003974>
- Smith, S. A., & O’Meara, B. C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. *Bioinformatics*, 28(20), 2689–2690. <https://doi.org/10.1093/bioinformatics/bts492>

