---
title: "Tutorial 2: Phylogenetics Pipeline: Inference & Dating"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Phylogenetics Pipeline: Inference & Dating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 8,
  fig.align = "center"
)
```

## Abstract

Phylogenetic inference represents the transition from a curated molecular dataset to an explicit evolutionary hypothesis. After assembling and preparing multilocus sequence matrices in [Tutorial 1](https://beeamerino.github.io/PhyloCactus/articles/tutorial-1-cactus-phylogeny-prep.html), the next challenge is to estimate evolutionary relationships while accounting for the heterogeneous processes that shape molecular evolution.

This tutorial describes the second stage of the `PhyloCactus` workflow, where concatenated multilocus datasets are transformed into statistically evaluated phylogenetic hypotheses. The pipeline integrates partition specific substitution model selection, constrained maximum likelihood inference, assessment of topological uncertainty, and divergence time estimation using penalized likelihood.

Large multilocus phylogenetic datasets often contain substantial uncertainty due to incomplete molecular sampling, uneven taxonomic representation, and heterogeneous sequence availability. However, relationships among major taxonomic lineages may already be supported by previous phylogenetic studies. To incorporate this available biological knowledge while allowing the placement of newly sampled species, `PhyloCactus` uses a taxonomic constraint scaffold defined in `cactus_constraints.csv`. This file assigns sampled species to established higher level clades and subfamilies, providing a biologically informed framework that reduces exploration of unsupported regions of tree space while preserving inference among unresolved relationships within major lineages.

Divergence time estimation requires the integration of molecular branch lengths with temporal information derived from fossil evidence or previously estimated evolutionary frameworks. Because **Cactaceae** lacks reliable direct fossil calibrations suitable for dating the entire family, `PhyloCactus` implements temporal constraints through the `calibrations_bounds.csv` file. These bounds incorporate lineage specific temporal estimates from Hernández-Hernández *et al*. (2014) for major cactus clades and subfamilies, together with secondary calibration information for **Cactaceae** based on the broader flowering plant framework proposed by Ramírez-Barahona *et al*. (2020). This strategy allows the estimation of a temporal framework while acknowledging the limitations imposed by the fossil record.

Because evolutionary history cannot be directly observed, phylogenetic reconstruction relies on statistical models that approximate the processes generating molecular variation. Therefore, the resulting phylogeny should be interpreted as the evolutionary hypothesis that best explains the observed molecular data under the selected models and assumptions, rather than as an absolute representation of historical relationships.

The workflow combines `ModelTest-NG`, `RAxML-NG`, and `treePL` to estimate a time calibrated phylogeny from the curated supermatrices generated in the previous tutorial. The resulting evolutionary framework provides the foundation for downstream comparative analyses, including historical biogeography, diversification dynamics, and trait evolution.

## Complete Pipeline Execution Workflow

### Setup: Creating a Clean Workspace

Before proceeding with **Module 7**, ensure that the required external phylogenetic binaries (`ModelTest-NG`, `RAxML-NG`, and `treePL`) are installed on your machine and accessible via your system's `$PATH`.

*If you are using macOS, many of these can be installed via Homebrew (`brew install raxml-ng`, etc.). On Linux/Windows, ensure they are compiled and mapped correctly to your `$PATH`. Alternatively, if you receive a "command not found" error, you can provide the absolute path to the executable file (e.g., `raxml_path = "/usr/local/bin/raxml-ng"`) directly in the function arguments.*

To quickly run the analyses in this module, you can copy the entire executable script to your tutorial directory:

``` r
tutorial_dir <- "~/Desktop/PhyloCactus_Tutorial"
setwd(tutorial_dir)

# Copy the entire tutorial script for easy execution
file.copy(
  system.file("scripts", "tutorial-2-cactus-phylogeny-inference.R", package = "PhyloCactus"),
  file.path(tutorial_dir, "tutorial-2-cactus-phylogeny-inference.R")
)
```

### Module 7: Preprocess and Substitution Models

Phylogenetic datasets derived from multiple molecular markers often contain substantial evolutionary heterogeneity. Different loci may evolve under distinct substitution patterns, nucleotide compositions, and selective constraints. If these differences are ignored, a single evolutionary model applied across the entire dataset may inadequately describe the observed sequence variation and introduce biases during likelihood based phylogenetic inference.

Therefore, before estimating the phylogenetic relationships, the concatenated supermatrix must be represented within a statistical framework that accounts for variation among molecular partitions. Partition specific substitution models provide a more realistic approximation of sequence evolution by allowing each locus region to be analyzed according to its own evolutionary characteristics.

The **`preprocess_partitions()`** function prepares the concatenated supermatrix generated in Module 6 for downstream phylogenetic inference. It reads the alignment and partition files, verifies their structural compatibility using the `RAxML-NG` parser, and generates standardized input files required for substitution model evaluation and maximum likelihood analysis.

The **`run_modeltest_ng()`** function performs automated substitution model evaluation for each predefined molecular partition. By interfacing with `ModelTest-NG` (Darriba *et al*., 2020; Flouri *et al*., 2015), the function compares alternative models of nucleotide evolution and identifies the model that provides the best statistical explanation of the observed sequence variation according to information criteria such as AICc (Akaike, 1974; Hurvich & Tsai, 1989). Selecting optimal partition-specific substitution models is crucial for controlling mutational rate heterogeneity across loci and mitigating systematic artifacts such as Long-Branch Attraction (LBA), where rapidly evolving non-homologous lineages false-group together. The resulting model scheme provides the statistical foundation for maximum-likelihood tree inference.

```{r, eval = FALSE}
library(PhyloCactus)

# Setup directories
output_dir <- "7_Phylogenetics"
dir.create(output_dir, showWarnings = FALSE)

# Get model test path. Note: Configure these in your .Renviron file
modeltest_path <- Sys.getenv("PATH_MODELTEST_NG", "modeltest-ng")
raxml_path     <- Sys.getenv("PATH_RAXML_NG", "raxml-ng")
treepl_path    <- Sys.getenv("PATH_TREEPL", "treePL")

preprocess_partitions(
  phy_matrix = "6_Concatenated/concatenated_alignments/ALIGNMENT_supermatrix.phy",
  part_file = "6_Concatenated/concatenated_alignments/PARTITION_iqtree.part",
  raxml_path = raxml_path,
  output_dir = output_dir
)

# Run `ModelTest-NG` to select substitution models for each partition
# Use alternative working directory to avoid modeltest error.
setwd(output_dir)
best_models <- run_modeltest_ng(
  modeltest_exec_path = modeltest_path,
  aln_file = "cactus_check.raxml.reduced.phy",
  part_file = "cactus_check.raxml.reduced.clean.partition",
  prefix = "cactus_phylo_modeltest",
  threads = 8
)
setwd("..")
```

### Module 8: Constraint Trees and Maximum Likelihood Search

Maximum likelihood inference estimates the evolutionary hypothesis that provides the best explanation of the observed molecular data under a defined substitution model. The resulting phylogeny should not be interpreted as an absolute representation of biological history, but rather as the topology that maximizes the likelihood of observing the molecular alignment given the evolutionary models, constraints, and available molecular evidence.

Large phylogenetic datasets frequently include incomplete taxonomic sampling, uneven sequence availability, and uncertainty associated with the placement of newly incorporated species. These challenges become particularly relevant when expanding datasets to include a large number of taxa from highly diverse evolutionary radiations. In such cases, previously established phylogenetic knowledge can be incorporated to reduce uncertainty and improve the efficiency of tree space exploration.

The **`build_constraint_scaffold()`** function generates a topological constraint scaffold from the taxonomic framework provided in `cactus_constraints.csv`. This file assigns sampled species to major evolutionary groups within **Cactaceae**, including recognized subfamilies and internal clades. These assignments are based on previously published phylogenetic evidence and current evolutionary consensus regarding relationships among major cactus lineages, particularly the framework established by Guerrero *et al*. (2019).

The constraint scaffold is not intended to replace molecular inference or impose a complete predefined phylogeny. Instead, it incorporates well supported relationships among higher level lineages while allowing unresolved relationships, particularly among newly incorporated species, to be inferred directly from the molecular dataset. By constraining only relationships supported by previous evidence, the approach reduces uncertainty during maximum likelihood inference and focuses the analysis on the evolutionary relationships that remain to be resolved.

The **`calculate_ml_tree()`** function interfaces with `RAxML-NG` (Kozlov *et al*., 2019) to estimate the maximum likelihood topology from the concatenated molecular matrix using the partition specific substitution models identified in **Module 7**. The analysis searches for the tree configuration that maximizes the likelihood of observing the sequence data under the selected evolutionary framework while incorporating the constraint scaffold defined for **Cactaceae**.

The resulting maximum likelihood tree represents the best supported evolutionary hypothesis under the available molecular evidence, substitution models, and prior phylogenetic framework. This topology serves as the reference tree for subsequent branch support estimation and divergence time analyses.

The **`calculate_rf_distances()`** function evaluates topological variability among independently inferred trees by calculating pairwise Robinson Foulds distances (Robinson & Foulds, 1981). These comparisons provide an additional measure of consistency among alternative tree searches and allow the assessment of whether independent analyses converge toward similar evolutionary hypotheses.

```{r, eval = FALSE}
# Assembles taxonomic classifications into a constraint scaffold
constraint_tree <- build_constraint_scaffold(
  alignment_path = file.path(output_dir, "cactus_check.raxml.reduced.phy"),
  constraints_csv_path = system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus"),
  output_dir = output_dir
)

# Execute maximum-likelihood search with `RAxML-NG`
ml_results <- calculate_ml_tree(
  raxml_bin_path = raxml_path,
  aln_file = file.path(output_dir, "cactus_check.raxml.reduced.phy"),
  part_file = file.path(output_dir, "cactus_phylo_modeltest.part.aicc"),
  constraint_file = file.path(output_dir, "cactus_constraints.tree"),
  outgroup = "Portulaca_fulgens",
  n_init_trees = "rand{25},pars{25}",
  seed = 1111,
  n_workers = 1,
  threads = 8, 
  output_dir = output_dir,
  prefix = "cactus_search"
)

# Calculate Robinson-Foulds distance among maximum-likelihood trees
cat("Calculating distances among maximum-likelihood trees...\n")
rf_dist <- calculate_rf_distances(
  raxml_bin_path = raxml_path,
  ml_trees_file = ml_results$mlTrees,
  output_dir = output_dir
)
```

### Module 9: Topological Validation via Bootstrap Support

Once the maximum likelihood topology has been inferred, the statistical support of recovered clades must be evaluated by quantifying topological stability associated with each internal node. Bootstrap analyses provide an empirical measure of topological stability by assessing how consistently specific relationships are recovered under resampling of the original sequence alignment.

The **`generate_bootstrap_script()`** and **`run_local_bootstraps()`** functions configure and execute bootstrap searches using `RAxML-NG`. These procedures generate independent resampled datasets while maintaining the same substitution model, constraint tree and partition scheme defined during the maximum likelihood inference. The resulting bootstrap trees represent alternative evolutionary hypotheses derived from the molecular dataset and allow the assessment of uncertainty across the inferred topology.

The **`collect_bootstraps()`** function gathers the independent bootstrap searches, while **`check_bs_convergence()`** evaluates whether a sufficient number of replicates has been generated using the bootstopping criterion (Pattengale *et al*., 2010). This procedure prevents unnecessary computational effort while ensuring that bootstrap support values have reached stable estimates.

The **`map_branch_supports()`** function summarizes bootstrap information and transfers support values onto the best maximum likelihood topology. Instead of relying exclusively on traditional bipartition frequencies, `PhyloCactus` uses Transfer Bootstrap Expectation (`TBE`; Lemoine *et al*., 2018), which provides a more informative measure of node support when analysing large phylogenies containing incomplete taxonomic sampling and heterogeneous amounts of missing data.

#### Temporal Bootstrap Replicates for Divergence Time Estimation

Divergence time estimation requires an additional assessment of uncertainty because branch lengths inferred during maximum likelihood analysis are influenced by sampling variation in the molecular data. Following the empirical protocol proposed by Maurin (2020) for dating large phylogenies with `treePL`, `PhyloCactus` generates a second set of bootstrap replicates specifically designed for temporal analyses.

The **`calculate_temporal_bootstraps()`** function generates bootstrap alignments and performs maximum likelihood searches while constraining the topology to the previously inferred maximum likelihood tree. This constraint does not force branch lengths to be identical; instead, it preserves the supported evolutionary relationships while allowing each bootstrap replicate to independently estimate branch length variation from the resampled molecular data.

This strategy is particularly important for large phylogenetic datasets where many relationships are already strongly supported at higher taxonomic levels, but uncertainty remains associated with branch length estimation and the placement of newly incorporated taxa. By maintaining a common topological framework, temporal bootstrap replicates provide a distribution of alternative branch length estimates that can be propagated into downstream divergence time analyses.

The resulting bootstrap trees are subsequently used by the dating module with `treePL` (Smith & O’Meara, 2012). Following the protocol described by Maurin (2020), the best maximum likelihood topology is first optimized to determine the appropriate penalized likelihood parameters and smoothing value. These parameters are then applied to the temporal bootstrap replicates to estimate a distribution of dated trees, from which confidence intervals for node ages can be obtained.

Therefore, `PhyloCactus` distinguishes between two complementary bootstrap procedures: standard bootstrap analyses evaluate the topological stability of inferred relationships, whereas temporal bootstrap replicates quantify uncertainty in molecular branch length estimates required for divergence time estimation.

```{r, eval = FALSE}
# -------------------------------------------------------------
# Estimate Bootstrap Replicates
# -------------------------------------------------------------
# `RAxML-NG` supports running Bootstraps either locally or via HPC array chunks.
# You can set run_local_bs to TRUE to run them now on your machine, 
# or FALSE to generate an HPC bash script to run them remotely.
run_local_bs <- FALSE

if (run_local_bs) {
  cat("\n1. Estimating bootstraps locally...\n")
  bs_dir <- file.path(output_dir, "local_bs")
  local_bs <- run_local_bootstraps(
    raxml_bin_path = raxml_path,
    aln_file = file.path(output_dir, "cactus_check.raxml.reduced.phy"),
    part_file = file.path(output_dir, "cactus_phylo_modeltest.part.aicc"),
    constraint_file = file.path(output_dir, "cactus_constraints.tree"),
    bs_trees = 500,
    outgroup = "Portulaca_fulgens",
    threads = 8,
    workers = 1,
    output_dir = bs_dir
  )
} else {
  cat("\n1. Generating BS script for HPC chunks...\n")
  bs_dir <- file.path(output_dir, "bs_chunks")
  bs_script <- generate_bootstrap_script(
    alignment_file = file.path(output_dir, "cactus_check.raxml.reduced.phy"),
    partition_file = file.path(output_dir, "cactus_phylo_modeltest.part.aicc"),
    constraint_file = file.path(output_dir, "cactus_constraints.tree"),
    outgroup = "Portulaca_fulgens",
    bs_per_rep = 500,
    max_reps = 2,   # E.g. to reach 1000 replicates using 2 scripts/jobs 
    threads = 120,
    workers = 20,
    output_dir = file.path(output_dir)
  )
}

# -------------------------------------------------------------
# Collect Bootstrap Replicates
# -------------------------------------------------------------
# Before running this step, ensure that your bootstrap replicates have finished.
# If you ran them on HPC, make sure all chunks completed successfully and
# optionally sync the files back to your local repository.

cat("\n2. Collecting bootstrap replicates...\n")
# bs_dir was defined in Step 1 depending on whether you ran locally or via HPC chunks
bs_concat_file <- collect_bootstraps(
  bs_dir = bs_dir,
  output_dir = output_dir
)
cat("   Bootstraps concatenated to:", bs_concat_file, "\n")

# -------------------------------------------------------------
# Check Bootstrap Convergence
# -------------------------------------------------------------
cat("\n3. Checking bootstrap convergence...\n")
converge_log <- check_bs_convergence(
  raxml_bin_path = raxml_path,
  bs_trees_file = file.path(output_dir, "cactus_ALL_bootstraps.tree"),
  bs_cutoff = 0.03,
  seed = 111,
  threads = 8,
  output_dir = output_dir,
  prefix = "cactus_bs_convergence"
)
cat("   Convergence log saved to:", converge_log, "\n")

# -------------------------------------------------------------
# Map TBE Supports onto Best Tree
# -------------------------------------------------------------
cat("\nMapping TBE supports onto the best tree...\n")

support_tree <- map_branch_supports(
  raxml_bin = raxml_path,
  best_tree = file.path(output_dir, "cactus_search.raxml.bestTree"),
  bootstraps_file = file.path(output_dir, "cactus_ALL_bootstraps.tree"),
  metric = "tbe",
  output_dir = output_dir,
  prefix = "cactus_support"
)
cat("   Support tree generated at:", support_tree, "\n")

# Note: for Dating, run sequential temporal bootstraps constrained over the best tree.
cat("Temporal Bootstraps for date intervals...\n")
temporal_bs <- calculate_temporal_bootstraps(
  raxml_bin_path = raxml_path,
  aln_file = file.path(output_dir, "cactus_check.raxml.reduced.phy"),
  part_file = file.path(output_dir, "cactus_phylo_modeltest.part.aicc"),
  best_tree_file = file.path(output_dir, "cactus_search.raxml.bestTree"),
  outgroup = "Portulaca_fulgens",
  bs_trees = 100, # For demo speed. Adjust to 500 or 1000.
  threads = 8,
  output_dir = file.path(output_dir, "cactus_temporal_bs")
)
```

### Module 10: Estimating Divergence Times via Penalized Likelihood

The inferred maximum likelihood topology provides a robust hypothesis of evolutionary relationships; however, branch lengths estimated from molecular data represent substitutions per site rather than geological time. To reconstruct the temporal history of **Cactaceae** diversification, the phylogenetic tree must therefore be transformed into an ultrametric chronogram by integrating molecular branch length information with temporal constraints derived from previous evolutionary studies.

`PhyloCactus` estimates divergence times using the penalized likelihood framework implemented in `treePL` (Sanderson, 2002; Smith & O’Meara, 2012). This approach is particularly suitable for large phylogenetic datasets because it accommodates lineage specific variation in evolutionary rates while remaining computationally feasible for analyses involving hundreds or thousands of taxa.

Temporal constraints are provided through the `calibrations_bounds.csv` reference file distributed with the package. This file defines minimum and maximum age boundaries for major **Cactaceae** lineages, including subfamilies and internal clades. Each calibration is assigned to specific nodes of the phylogeny based on taxonomic relationships and previously published divergence time estimates.

The calibration strategy follows the current understanding of **Cactaceae** evolutionary history. Internal calibrations for major cactus lineages are derived from the temporal framework proposed by Hernández-Hernández *et al.* (2014), which estimated divergence times among the principal evolutionary groups of the family, including **Opuntioideae**, **Cactoideae**, and major internal clades within these lineages.

Because **Cactaceae** lacks a sufficiently informative fossil record suitable for direct calibration of the family crown age, the temporal constraint for the family level node is derived from a secondary calibration based on the dated phylogenetic framework of Ramírez-Barahona *et al.* (2020). This strategy follows common practice in large scale plant phylogenies, where broader temporal constraints derived from independent dated analyses are used when direct fossil evidence is unavailable or insufficient.

The use of secondary calibration for the family crown age allows the analysis to incorporate current evolutionary knowledge while explicitly acknowledging uncertainty associated with the absence of direct fossil evidence. Rather than relying on uncertain fossil assignments, `PhyloCactus` applies transparent and reproducible temporal boundaries derived from previous comprehensive studies.

The **`automate_treePL()`** function performs the complete divergence time estimation workflow. First, it prepares the maximum likelihood topology, branch length information, and calibration constraints required by `treePL`. The function then performs the optimization procedure required to estimate appropriate penalized likelihood parameters, including the smoothing parameter selected through cross validation.

Following the empirical protocol described by Maurin (2020), the optimization and cross validation steps are performed using the best scoring maximum likelihood tree without bootstrap annotations. This approach ensures that the dating parameters are optimized using the primary evolutionary hypothesis before being propagated to replicate datasets.

After the optimal `treePL` parameters have been identified, the dating procedure is applied to the temporal bootstrap replicates generated in **Module 9**. These replicates were obtained by resampling the original alignment while maintaining the maximum likelihood topology as a constraint. This strategy preserves the inferred relationships among taxa while allowing branch lengths to vary according to molecular sampling uncertainty.

Propagating temporal uncertainty through these bootstrap replicates allows divergence time estimates to incorporate variation in branch length estimation. The resulting collection of dated bootstrap trees represents uncertainty associated with the temporal reconstruction under a maximum likelihood framework and provides the basis for estimating confidence intervals around divergence times.

#### Summarizing Dated Temporal Bootstrap Replicates

After all temporal bootstrap replicates have been independently dated, the resulting chronograms must be summarized into a single representative time calibrated tree. Following the empirical workflow proposed by Maurin (2020), `PhyloCactus` uses `TreeAnnotator` (Helfrich *et al.*, 2018) to summarize the dated bootstrap distribution.

Because temporal bootstrap replicates are generated under the maximum likelihood topology constraint, all replicate trees share identical branching relationships. Therefore, the summary procedure does not evaluate alternative topologies, but instead estimates the distribution of node ages across independently dated replicate trees.

The recommended `TreeAnnotator` configuration specifies a maximum sum of clade credibility target tree type, mean height node age estimates, and 0% burn-in.

Under this framework, `TreeAnnotator` calculates the mean age of each node and summarizes the uncertainty associated with divergence time estimation across the dated temporal bootstrap replicates.

TreeAnnotator reports these uncertainty intervals as highest posterior density (HPD) values due to compatibility with Bayesian phylogenetic workflows. However, these values should not be interpreted as Bayesian posterior distributions. Within the `PhyloCactus` framework, they represent confidence intervals derived from variation among independently dated temporal bootstrap replicates generated through penalized likelihood optimization.

Here is how you execute, print, and visualize these chronological calculations:

```{r, eval = FALSE}
dating_dir <- "8_Dating"
dir.create(dating_dir, showWarnings = FALSE)

# 1. Prepare Calibrations
cat("Creating `treePL` calibrations configurations...\n")
calibs_all <- read.csv(system.file("extdata", "calibrations_bounds.csv", package = "PhyloCactus"))
calibs <- calibs_all[calibs_all$used_in_analysis == TRUE, ]

constraints <- read.csv(system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus"))
ml_tree <- ape::read.tree(file.path(output_dir, "cactus_search.raxml.bestTree"))
tip_labels <- ml_tree$tip.label
num_sites <- 12700

cfg_lines <- c()
for (i in seq_len(nrow(calibs))) {
  row <- calibs[i, ]
  tips_all <- unique(constraints[constraints[[row$column]] == row$value, "Specie_name"])
  tips_in_tree <- tips_all[tips_all %in% tip_labels]
  
  if (length(tips_in_tree) < 2) next
  
  mrca_line <- paste("mrca =", row$mrca, paste(tips_in_tree, collapse = " "))
  min_line  <- sprintf("min = %s %f", row$mrca, row$min)
  max_line  <- sprintf("max = %s %f", row$mrca, row$max)
  
  cfg_lines <- c(cfg_lines, mrca_line, min_line, max_line)
}

treepl_cfg <- c(
  paste0("numsites = ", num_sites),
  cfg_lines,
  "nthreads = 8",
  "thorough"
)

calibrations_cfg_path <- file.path(dating_dir, "calibrations_treePL_fulltips.cfg")
writeLines(treepl_cfg, calibrations_cfg_path)
cat("   Calibrations compiled successfully to:", calibrations_cfg_path, "\n")

# 2. Run Fast Automated `treePL` Dating Pipeline
cat("\nRunning automated `treePL` wrapper script over maximum-likelihood tree and bootstrap replicates...\n")

# Provide the wrapper shell script provided in your extdata
# See reference: https://github.com/tongjial/treepl_wrapper
wrapper_sh_path <- system.file("extdata", "treepl_wrapper_v1.sh", package = "PhyloCactus")

# Note: For tutorial purposes, you can limit the number of bootstrap trees to process
# by setting `num_bs = 100` (or any other number). If not provided, it will process all
# available bootstrap trees. Here we set it to 100 for faster tutorial execution.
automate_treePL(
  cfg_file = calibrations_cfg_path,
  wrapper_sh = wrapper_sh_path,
  ml_tree_file = file.path(output_dir, "cactus_search.raxml.bestTree"),
  bs_trees_file = file.path(output_dir, "cactus_temporal_bs", "cactus_temporal_bs.raxml.bootstraps"),
  results_dir = file.path(dating_dir, "auto_results"),
  treePL_out = dating_dir,
  num_bs = 100
)

# 3. View and evaluate dating metrics cleanly in R
cat("\n--- Chronological Dating Results Summary ---\n")
if (file.exists(file.path(dating_dir, "BestTree_treePL.tree"))) {
  ml_chronogram <- ape::read.tree(file.path(dating_dir, "BestTree_treePL.tree"))
  cat("   Best ML Chronogram:\n")
  cat("     - File path:", file.path(dating_dir, "BestTree_treePL.tree"), "\n")
  cat("     - Root age:", max(ape::node.depth.edgelength(ml_chronogram)), "Mya\n")
  cat("     - Number of tips:", length(ml_chronogram$tip.label), "\n")
}
cat("--------------------------------------------\n")
```

## Next Steps

At this stage, phylogenetic inference and divergence time estimation have been completed. The resulting maximum likelihood topology, branch support estimates, and time calibrated chronogram provide the evolutionary framework required for downstream comparative analyses.

The next stage of the `PhyloCactus` workflow focuses on integrating phylogenetic outputs with biological metadata and generating publication quality visualizations. In the following tutorial, users will learn how to visualize the maximum likelihood phylogeny with branch support values, display the dated chronogram with temporal uncertainty estimates, and integrate external biodiversity information, including IUCN Red List conservation categories.

[Continue to Tutorial 3: Data Visualization and IUCN Summaries](https://beeamerino.github.io/PhyloCactus/articles/tutorial-3-cactus-phylogeny-visualization.html)

## References

- Akaike, H. 1974. A new look at the statistical model identification. *IEEE Transactions on Automatic Control*, 19(6), 716-723. <https://doi.org/10.1109/TAC.1974.1100705>
- Arakaki *et al*. 2011. Contemporaneous and recent radiations of the world’s major succulent plant lineages. *Proceedings of the National Academy of Sciences of the United States of America*, 108(20), 8379–8384. <https://doi.org/10.1073/pnas.1100628108>
- Darriba *et al*. 2020. ModelTest-NG: a new and scalable tool for the selection of DNA and protein evolutionary models. *Molecular Biology and Evolution*, 37(1), 291-294. <https://doi.org/10.1093/molbev/msz189>
- Flouri *et al*. 2014. The Phylogenetic Likelihood Library. *Systematic Biology*, 64(2), 356-362. <https://doi.org/10.1093/sysbio/syu084>
- Helfrich *et al*. 2018. TreeAnnotator: versatile visual annotation of hierarchical text relations. *Proceedings of the Eleventh International Conference on Language Resources and Evaluation*. <https://lrec.elra.info/lrec2018-main-308>
- Hernández-Hernández *et al*. 2014. Beyond aridification: Multiple explanations for the elevated diversification of cacti in the New World Succulent Biome. *New Phytologist*, 202(4), 1382–1397. <https://doi.org/10.1111/nph.12752>
- Kozlov *et al*. 2019. RAxML-NG: A fast, scalable and user-friendly tool for maximum likelihood phylogenetic inference. *Bioinformatics*, 35(21), 4453–4455. <https://doi.org/10.1093/bioinformatics/btz305>
- Lemoine *et al*. 2018. Renewing Felsenstein’s phylogenetic bootstrap in the era of big data. *Nature*, 556(7702), 452-456. <https://doi.org/10.1038/s41586-018-0043-0>
- Maurin, K. J. 2020. An empirical guide for producing a dated phylogeny with treePL in a maximum likelihood framework. *arXiv preprint arXiv:2008.07054*. <https://doi.org/10.48550/arXiv.2008.07054>
- Pattengale *et al*. 2010. How many bootstrap replicates are necessary?. *Journal of Computational Biology*, 17(3), 337–354. <https://doi.org/10.1089/cmb.2009.0179>
- Ramírez-Barahona *et al*. 2020. The delayed and geographically heterogeneous diversification of flowering plant families. *Nature Ecology and Evolution*, 4(9), 1232–1238. <https://doi.org/10.1038/s41559-020-1241-3>
- Robinson, D. F., & Foulds, L. R. 1981. Comparison of phylogenetic trees. *Mathematical Biosciences*, 53(1-2), 131-147. <https://doi.org/10.1016/0025-5564(81)90043-2>
- Sanderson, M. J. 2002. Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach. *Molecular Biology and Evolution*, 19(1), 101–109. <https://doi.org/10.1093/oxfordjournals.molbev.a003974>
- Smith, S. A., & O’Meara, B. C. 2012. TreePL: Divergence time estimation using penalized likelihood for large phylogenies. *Bioinformatics*, 28(20), 2689–2690. <https://doi.org/10.1093/bioinformatics/bts492>

