---
title: "Tutorial 5: DNA Barcoding, Section Manifest"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 5: DNA Barcoding, Section Manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

This section of `PhyloCactus` is under construction and contains no implementation.
This document sets out what the molecular diagnostic section is intended to do, the design decisions already taken, the questions that remain open, and the conditions for implementing it.

## Scope

The section is intended to answer one question: given a curated multilocus reference library for Cactaceae, how reliably can a single locus, or a combination of loci, assign an unknown sample to genus and to species, and in which groups does that assignment fail?

The planned components are:

1. **Empirical barcode gap per locus.** Intraspecific and interspecific pairwise distance distributions for each marker, with the gap reported as an observed quantity and not tested against a fixed cutoff.
2. **Probabilistic classification with cross-validation.** A taxonomic classifier trained on sequences of accepted species and evaluated by cross-validation stratified by species. Species represented by a single accession cannot be evaluated by leaving one sequence out, and are reported separately.
3. **Monophyly in single-locus gene trees.** Recovery of species and genus monophyly for each marker. Non-monophyly is interpreted in terms of incomplete lineage sorting, hybridization and plastid capture, and paraphyletic assemblages are reported as grades.

## Decisions Taken

- **Taxonomic backbone.** Names are reconciled against the Cactaceae checklist of Caryophyllales.org (Korotkova *et al*. 2021), the same backbone used in the assembly modules.
- **Input.** The section uses the curated ingroup alignments in `4_Cleaned/cleaned_markers_ingroup` (Module 4) and the phylogenetic products of Modules 7 to 10. It does not query GenBank again.
- **No fixed distance thresholds.** Any decision rule must be derived from the observed distance distributions of the family, and its derivation must be reported with the result. In a family with recent radiations and low plastid divergence, a fixed threshold assumes a level of divergence that is not uniform across genera.
- **Validation.** Classification accuracy is reported only from cross-validation stratified by species. Accuracy measured on the training sequences reflects how well the model fits those sequences, not its performance on new samples.

## Open Questions

1. **Loci and combinations.** Whether resolution is assessed per locus, on concatenated subsets, or both, and whether the standard plant barcode combination is evaluated as a unit for comparison with published surveys.
2. **Unit of assignment.** Whether the target is assignment to species, assignment to genus with an explicit statement of ambiguity at species level, or a graded output. Many plant species are not monophyletic in single-locus trees, so the output must allow for justified non-assignment.
3. **Reporting ambiguity.** Whether a sample that cannot be separated from a set of candidates is reported as a candidate set, as an assignment to genus, or as a failure, and how this affects the conservation applications of the package.
4. **Genome skims and other reads.** Decided (BMM, 2026-10-07): the library is built from Sanger marker records of GenBank and own sequences, and genome skims are not a source of it. Illumina reads that a user brings (skim or whole-genome sequencing) are a query: `identify_barcoding_reads()` assembles their plastome and nrDNA and identifies them against the library, locus by locus.
5. **Paralogy and organelle capture.** Whether loci with cytonuclear conflict are excluded from diagnostic use, flagged, or reported as evidence of reticulation.

## How Well This Version Identifies

The figures below were measured on the library of this development version (frozen on 2026-10-06; manifest md5 `48be326e9ce4f1b779a175e149d509d1`), with the species thresholds of step 9 (quantile 0.95 of the species confidence of the genus scheme, floor 60).
They belong to this library: as GenBank and own sequences enrich it in later versions, the probes that produced them are rerun and the rates are expected to change.

Outcomes of real queries, in the order species correct / genus correct / abstention / conflict / wrong species / wrong genus.
A species is "present" when the library holds it once the records of the query's own specimen are set aside.

| Input | Unit | Species present | Species absent |
|---|---|---|---|
| One Sanger record | locus | 168 / 405 / 1733 / n.a. / 15 / 4 (2325) | 0 / 286 / 2181 / n.a. / 48 / 20 (2535) |
| Sanger records of one GenBank specimen | specimen | 6 / 52 / 101 / 0 / 3 / 0 (162) | 0 / 20 / 17 / 0 / 1 / 0 (38) |
| Assembled plastome from GenBank | plastome | 5 / 51 / 30 / 0 / 2 / 5 (93) | 0 / 2 / 0 / 1 / 0 / 1 (4) |
| Illumina reads from SRA | run | 1 / 1 / 2 / 0 / 0 / 0 (4) | 0 / 1 / 2 / 0 / 0 / 0 (3) |

For single records and sets of records, the records of the query's specimen (and, for single records, of the other queries identified with it) were removed from the library before identification.
The single-record rows include 55 own records among 4860.
Plastomes and reads were identified against the full library.

Abstention is the most common answer of this version, and a wrong species is rare but possible.
Six public examples, one per kind of answer, show why:

| Answer | Input | Accession | True species | What the package answered |
|---|---|---|---|---|
| Species correct | plastome | MN517611.1 | *Mammillaria zephyranthoides* (in the library) | the species from `matK` (confidence 80.7, threshold 67.4), among 114 species of *Mammillaria* in the `matK` library |
| Genus correct | reads | SRR26398994 | *Opuntia comonduensis* (not in the library) | *Opuntia* from `ndhF-rpl32` (genus confidence 80.8); the species confidence stays low among 71 congeners |
| Abstention | reads | SRR12935164 | *Ariocarpus retusus* (in the library) | no answer: 13 loci assembled, all below 60; the best rows point to *Ariocarpus* (48.7 and 48.0) |
| Conflict | plastome | OQ458704.1 | *Cleistocactus hyalacanthus* (not in the library) | *Cleistocactus* from `trnS-trnG` against *Cereus* from `psbJ-petA`, whose library holds no *Cleistocactus* |
| Wrong species | plastome | MT387452.1 | *Rhipsalis teres* (in the library) | *Lepismium lumbricoides* from `matK`: the region is identical to FN669720.1, a record under that name |
| Wrong genus | plastome | OR807739.1 | *Parodia microsperma* (in the `matK` library) | *Eriosyce* from `rbcL`, whose library holds no *Parodia* |

- A species is named only when one locus clears the species threshold of that locus.
- A species absent from the library is answered at genus rank when its congeners keep the species confidence low.
- A locus whose library lacks the true genus can only name another genus; when other loci disagree, the answer is a conflict, not a choice. The coverage of each locus library tells which genera it holds.
- An identical sequence under two names (a misidentified voucher or a haplotype shared across genera) yields a confident wrong answer. The report lists the identical library record so that it can be inspected.

The same limitation weighs on the Chilean genera: in the single-record probe, the `matK` record MT792775.1 of *Eulychnia acida*, with its species absent from the reduced library, was named *E. iquiquensis*, the only congener left (confidence 77.9, threshold 67.4).

The examples are reproduced from the tutorial folder, with the library of step 4 and the thresholds of step 9:

```{r, eval = FALSE}
library(PhyloCactus)

# A plastome downloaded from GenBank as a flat file
identify_barcoding_query("MN517611.1.gb", run_name = "MN517611.1")

# Illumina reads downloaded from ENA (fastp and GetOrganelle required)
identify_barcoding_reads("SRR26398994_1.fastq.gz", "SRR26398994_2.fastq.gz",
                         run_name = "SRR26398994")
```

## Conditions for Implementation

Two conditions are met in version 0.4.5: a dated and supported phylogeny is available (Modules 7 to 10), so monophyly per locus can be assessed against a reference topology, and the metadata and visualization modules are available (Modules 11 and 12), so diagnostic results can be related to conservation status and distribution.
Two conditions remain:

- the five open questions above are answered and recorded;
- a validation plan is written before any code, specifying the cross-validation scheme, the treatment of species with a single accession, and the negative controls.

[Continue to Tutorial 6: Core Pipeline Functions and Methodological Dictionary](https://beeamerino.github.io/PhyloCactus/articles/tutorial-6-cactus-phylogeny-functions.html)

## References

- Korotkova *et al*. 2021. Cactaceae at Caryophyllales.org - A dynamic online species-level taxonomic backbone for the family. *Willdenowia*, 51(2), 251–270. <https://doi.org/10.3372/wi.51.51208>
