---
title: "Tutorial 4: Phylogenetic Validation and Comparative Analyses"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 4: Phylogenetic Validation and Comparative Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

This tutorial compares the tree inferred in Tutorials 1 and 2 with four published phylogenies of Cactaceae.
A supermatrix tree is one estimate of the phylogeny; taxon sampling, choice of loci, substitution models and inference method can all change the topology.
Comparison with independent trees shows which relationships are shared and which differ.

The reference trees are those of Amaral *et al*. (2022), Thompson *et al*. (2024), Zuntini *et al*. (2024) and de Vos *et al*. (2025); the last two are based on hundreds of nuclear genes.
The trees are compared with the Robinson-Foulds distance (Robinson & Foulds 1981) and the Different Phylogenetic Information distance (Smith 2020), and the distances are projected into tree space by multidimensional scaling (MDS).

## Complete Pipeline Execution Workflow

### Module 13: Comparison with Published Phylogenies

#### Reference Trees

The four reference trees are distributed with the package in `inst/extdata/`, so the comparison runs without downloads: Amaral *et al*. (2022), Thompson *et al*. (2024), Zuntini *et al*. (2024), and a species tree estimated with `ASTRAL-III` from the gene trees of de Vos *et al*. (2025).
Paths to external software and to data downloaded by the user are read from environment variables.

#### Species Tree from the Gene Trees of de Vos *et al*. (2025)

Gene trees estimated from multilocus and target-capture data often disagree.
Part of this discordance has biological causes, such as incomplete lineage sorting and hybridization, and under high levels of incomplete lineage sorting a concatenated analysis can be statistically inconsistent (Kubatko & Degnan 2007).

`ASTRAL-III` (Zhang *et al*. 2018) estimates a species tree from a set of unrooted gene trees under the multispecies coalescent, by finding the species tree that shares the largest number of quartets with the gene trees.
The `ASTRAL-III` Java archive (`.jar`) must be downloaded separately.

#### Paths in `.Renviron`

Store paths to local software and data in `~/.Renviron`.
Shared scripts then contain no local paths.

```{r, eval = FALSE}
# Add these lines to your ~/.Renviron file (e.g., using usethis::edit_r_environ())
# ASTRAL_PATH="/path/to/astral.5.7.8.jar"
# DEVOS_GENETREES_PATH="/path/to/QC.bestTreeCollapsed.trees"
# DEVOS_METADATA_PATH="/path/to/606_2025_1948_MOESM1_ESM.csv"

# Read them in R:
astral_jar <- Sys.getenv("ASTRAL_PATH")
gene_trees_file <- Sys.getenv("DEVOS_GENETREES_PATH")
metadata_file <- Sys.getenv("DEVOS_METADATA_PATH")
```

#### Running `ASTRAL-III`

de Vos *et al*. (2025) sequenced the Angiosperms353 loci.
The reference tree used here is the `ASTRAL-III` species tree estimated from their gene trees, after the tip labels are converted to species names with the sample metadata of the study.

The chunk below renames the tips of the gene trees, writes them to a file, and runs `ASTRAL-III`.
With `run_astral_locally <- FALSE`, the default, it copies the precomputed tree distributed with the package instead.
Output files are written to a temporary directory; change `val_dir` to keep them.

```{r}
# Target output paths in the validation directory
val_dir <- file.path(tempdir(), "10_Validation")
dir.create(val_dir, recursive = TRUE, showWarnings = FALSE)
output_tree <- file.path(val_dir, "QC.Species_tree_astral.tree")
renamed_gene_trees_file <- file.path(val_dir, "QC.Species_tree_astral_rename.tree")

# Set to TRUE to run ASTRAL-III; FALSE (default) copies the precomputed tree.
run_astral_locally <- FALSE

astral_jar       <- Sys.getenv("ASTRAL_PATH")
devos_gene_trees <- Sys.getenv("DEVOS_GENETREES_PATH")
devos_metadata   <- Sys.getenv("DEVOS_METADATA_PATH")

java_bin <- Sys.which("java")
if (!nzchar(java_bin) && file.exists("/opt/homebrew/opt/openjdk/bin/java")) {
  java_bin <- "/opt/homebrew/opt/openjdk/bin/java"
}

if (run_astral_locally && nzchar(astral_jar) && file.exists(astral_jar) && 
    nzchar(devos_gene_trees) && file.exists(devos_gene_trees) && 
    nzchar(devos_metadata) && file.exists(devos_metadata) && 
    nzchar(java_bin)) {
  
  meta <- read.csv(devos_metadata, stringsAsFactors = FALSE, check.names = FALSE)
  meta$species_name <- gsub(" ", "_", meta$`Scientific name`)
  meta$Sample_from_tree <- stringr::str_extract(meta$`Name in tree`, "P[0-9]+")
  
  map_sample <- setNames(meta$species_name, meta$Sample_from_tree)
  map_srr    <- setNames(meta$species_name, meta$`ENA run acc.`)
  
  gene_trees <- ape::read.tree(devos_gene_trees)
  
  rename_tips <- function(tree) {
    tips <- tree$tip.label
    paftol_to_sample <- function(x) {
      num <- stringr::str_extract(x, "[0-9]+")
      paste0("P", stringr::str_sub(num, -5))
    }
    sample_ids <- ifelse(stringr::str_detect(tips, "PAFTOL"), paftol_to_sample(tips), NA)
    species_vec <- ifelse(stringr::str_detect(tips, "PAFTOL"), map_sample[sample_ids], map_srr[tips])
    species_vec[is.na(species_vec)] <- tips[is.na(species_vec)]
    tree$tip.label <- species_vec
    return(tree)
  }
  
  gene_trees <- lapply(gene_trees, rename_tips)
  class(gene_trees) <- "multiPhylo"
  ape::write.tree(gene_trees, file = renamed_gene_trees_file)
  
  cmd <- paste(shQuote(java_bin), "-jar", shQuote(astral_jar), 
               "-i", shQuote(renamed_gene_trees_file), 
               "-o", shQuote(output_tree), "2>/dev/null")
  system(cmd)
} else {
  # Copy the precomputed de Vos et al. (2025) species tree distributed with the package
  ref_devos <- system.file("extdata", "reference_devos_2025_astral.tree", package = "PhyloCactus")
  file.copy(ref_devos, output_tree, overwrite = TRUE)
}
```

### Topological Comparison

The focal tree is the chronogram from Module 10, whose topology is that of the maximum-likelihood tree (all 1021 bipartitions shared).
`validate_phylogenies()` restricts every tree to Cactaceae and to the names accepted in the checklist, then computes pairwise distances on four taxon sets:

- **A1:** species shared by all trees;
- **A2:** species shared by the focal tree and the trees of Thompson *et al*. (2024) and Amaral *et al*. (2022);
- **A3:** all species in the genera shared by all trees;
- **A4:** one species per genus shared by all trees.

Two distances are computed for each pair of trees.
The normalized Robinson-Foulds distance (RF) is the proportion of bipartitions that are not shared by the two trees.
The Different Phylogenetic Information distance (DPI) also gives partial credit to bipartitions that are similar but not identical.
Both are computed on unrooted trees.
Each distance matrix is projected onto two dimensions by classical MDS; trees that lie close together have similar topologies.

**Rooting.** `validate_phylogenies()` roots every tree on *Leuenbergeria*.
The Module 10 chronogram is rooted on Talinaceae, but the family filter removes all terminals outside Cactaceae, and *Leuenbergeria* is sister to the rest of the family in the focal tree.
The genus can be changed with `rooting_genus`; trees without it are left unrooted and flagged in `SUPP_rooting_summary`.
Because RF and DPI are computed on unrooted trees, the rooting affects only the exported tree files, not the distances or the tree-space projection.

**Trees compared.** The four reference trees are read from `inst/extdata/`, where they are distributed with the package. The focal tree is the chronogram produced in Module 10, `8_Dating/dated_summary_hpd.tree`. When that file is absent, as when this tutorial is run on its own, the reference chronogram distributed with the package (`phylocactus_chronogram_hpd.tree`) is used instead.

```{r, fig.width=11.69, fig.height=8.27, warning=FALSE, message=FALSE}
library(PhyloCactus)

# The focal tree is the chronogram of this run; the reference trees are distributed with the package.
focal_tree <- file.path("8_Dating", "dated_summary_hpd.tree")
if (!file.exists(focal_tree)) {
  focal_tree <- system.file("extdata", "phylocactus_chronogram_hpd.tree", package = "PhyloCactus")
}
tree_paths <- list(
  FocalTree   = focal_tree,
  Zuntini     = system.file("extdata", "reference_zuntini_2024.tree", package = "PhyloCactus"),
  Thompson    = system.file("extdata", "reference_thompson_2024.tree", package = "PhyloCactus"),
  Amaral      = system.file("extdata", "reference_amaral_2022.tree", package = "PhyloCactus"),
  deVos       = system.file("extdata", "reference_devos_2025_astral.tree", package = "PhyloCactus")
)

# Drop trees whose files are missing
existing_trees <- tree_paths[sapply(tree_paths, function(x) file.exists(x) && nzchar(x))]

plot_to_show <- NULL
table_to_show <- NULL

if(length(existing_trees) > 1) {
  checklist_file   <- system.file("extdata", "CactaceaeFullList_2026_07_01_Beatriz_Merino.xlsx", package = "PhyloCactus")
  constraints_file <- system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus")
  
  # validate_phylogenies() harmonizes tip names, computes RF and DPI distances,
  # and projects them into tree space.
  validation_outputs <- validate_phylogenies(
    trees_mapping_list = existing_trees,
    checklist_csv = checklist_file,
    constraints_map = constraints_file,
    out_dir = file.path(tempdir(), "10_Validation")
  )
  
  plot_to_show <- validation_outputs$plot
  
  dist_table <- read.csv(file.path(validation_outputs$tables, "TABLE_pairwise_tree_distances_main.csv"))
  table_to_show <- head(dist_table)
} else {
  message("Please upload the reference trees to inst/extdata/ to run the full validation.")
}

plot_to_show
```

```{r, echo=FALSE}
if (!is.null(table_to_show)) {
  knitr::kable(table_to_show, caption = "Pairwise tree distances")
}
```

`TABLE_pairwise_tree_distances_main.csv` lists the distances for every pair of trees and every taxon set, and `FIG_tree_space_main.pdf` shows the tree-space projections.
The distances summarize overall agreement between topologies; they do not identify which clades are in conflict.

## Conclusion

This tutorial completes the core workflow of `PhyloCactus`: assembly of a multilocus supermatrix from GenBank, maximum-likelihood inference with bootstrap support, divergence time estimation with penalized likelihood, integration of IUCN Red List data, figures, and comparison with published phylogenies.

The next section of the workflow will assess how well individual markers discriminate species.
It is under construction and contains no implementation.
Tutorial 5 describes its scope, the design decisions taken, the open questions, and the conditions for reopening it.

[Continue to Tutorial 5: DNA Barcoding, Section Manifest](https://beeamerino.github.io/PhyloCactus/articles/tutorial-5-cactus-phylogeny-barcoding.html)

## References

- Amaral *et al*. 2022. Spatial patterns of evolutionary diversity in Cactaceae show low ecological representation within protected areas. *Biological Conservation*, 273, 109677. <https://doi.org/10.1016/j.biocon.2022.109677>
- de Vos *et al*. 2025. Phylogenomics and classification of Cactaceae based on hundreds of nuclear genes. *Plant Systematics and Evolution*, 311(5), 28. <https://doi.org/10.1007/s00606-025-01948-z>
- Kubatko & Degnan 2007. Inconsistency of phylogenetic estimates from concatenated data under coalescence. *Systematic Biology*, 56(1), 17–24. <https://doi.org/10.1080/10635150601146041>
- Robinson & Foulds 1981. Comparison of phylogenetic trees. *Mathematical Biosciences*, 53(1–2), 131–147. <https://doi.org/10.1016/0025-5564(81)90043-2>
- Smith 2020. Information theoretic generalized Robinson–Foulds metrics for comparing phylogenetic trees. *Bioinformatics*, 36(20), 5007–5013. <https://doi.org/10.1093/bioinformatics/btaa614>
- Thompson *et al*. 2024. Identifying the multiple drivers of cactus diversification. *Nature Communications*, 15(1), 7114. <https://doi.org/10.1038/s41467-024-51666-2>
- Zhang *et al*. 2018. ASTRAL-III: polynomial time species tree reconstruction from partially resolved gene trees. *BMC Bioinformatics*, 19(Suppl 6), 153. <https://doi.org/10.1186/s12859-018-2129-y>
- Zuntini *et al*. 2024. Phylogenomics and the rise of the angiosperms. *Nature*, 629, 843–850. <https://doi.org/10.1038/s41586-024-07324-0>

