---
title: "Tutorial 3: Phylogenetics Pipeline: Data Visualization and IUCN Summaries"
author: "Beatriz M. Meriño"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Tutorial 3: Phylogenetics Pipeline: Data Visualization and IUCN Summaries}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
tutorial_dir <- path.expand("~/Desktop/PhyloCactus_Tutorial")
```

## Abstract

This tutorial represents the final stage of the core phylogenetic reconstruction workflow implemented in **`PhyloCactus`**, focusing on the integration, visualization, and biological interpretation of the inferred evolutionary framework.
Following the assembly of the multilocus supermatrix, maximum-likelihood inference, branch support assessment, and divergence time estimation described in [Tutorial 1](https://beeamerino.github.io/PhyloCactus/articles/tutorial-1-cactus-phylogeny-prep.html) and [Tutorial 2](https://beeamerino.github.io/PhyloCactus/articles/tutorial-2-cactus-phylogeny-inference.html), this module transforms the resulting phylogenetic outputs into an integrative evolutionary resource.

In **Module 11**, phylogenetic results are enriched with external biological information by integrating species-level metadata and conservation assessments from the **IUCN Red List**.
This step generates a comprehensive evolutionary registry that links molecular sampling effort, taxonomic validation, distributional information, and conservation status.
These integrated data provide the foundation for evaluating biases in molecular representation, identifying threatened evolutionary lineages, and exploring patterns of conservation vulnerability across the cactus tree of life.

In **Module 12**, the final phylogenetic products and molecular datasets are visualized and synthesized for downstream comparative analyses.
The sequence dataset and concatenated supermatrix are evaluated for locus coverage and informativeness, while the maximum-likelihood topology and time-calibrated chronogram are displayed with branch support values (FBP) and divergence time uncertainty intervals.
Finally, taxonomic reconciliation and macroevolutionary patterns across IUCN conservation categories, endemism, and ecological threats are synthesized into publication-ready visual summaries.

Together, this tutorial completes the construction of a reproducible phylogenetic framework for **Cactaceae**, integrating evolutionary history, temporal diversification, and conservation information into a unified analytical resource suitable for macroevolutionary, biogeographic, and conservation studies.

## Complete Pipeline Execution Workflow

### Setup: Creating a Clean Workspace

Before proceeding, ensure you have the required packages installed and set up.
We rely on the `rredlist` package to interact securely with the IUCN Red List API.

To install the required dependencies:

``` r
required_pkgs <- c("dplyr", "tidyr", "stringr", "readr", "ggplot2", "forcats", "purrr", "scales", "rredlist")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[,"Package"])]
if(length(new_pkgs)) install.packages(new_pkgs)
```

**IUCN Red List API Access:** To retrieve threat categories automatically, generate a free API token from the [IUCN Red List API Token Generation Page](https://apiv3.iucnredlist.org/api/v3/token), submit the request form, and store the resulting key in your `.Renviron` file using `usethis::edit_r_environ()` under the variable `IUCN_REDLIST_KEY="your_api_key"`.

### Data Setup & Pre-processing

To quickly run the analyses in this module, you can copy the entire executable script to your tutorial directory:

``` r
tutorial_dir <- path.expand("~/Desktop/PhyloCactus_Tutorial")
setwd(tutorial_dir)

# Copy the entire tutorial script for easy execution
file.copy(
  system.file("scripts", "tutorial-3-cactus-phylogeny-visualization.R", package = "PhyloCactus"),
  file.path(tutorial_dir, "tutorial-3-cactus-phylogeny-visualization.R")
)
```

### Module 11: Taxonomic Reconciliation and IUCN Metadata Enrichment

This vignette assumes that you have completed [Tutorial 2: Inference & Dating](https://beeamerino.github.io/PhyloCactus/articles/tutorial-2-cactus-phylogeny-inference.html).
Following the completion of maximum-likelihood phylogenetic inference, branch support estimation, and divergence time reconstruction, this module focuses on enriching the resulting evolutionary framework with biological metadata and generating comprehensive visual summaries of the final multilocus dataset.

#### The Scientific Importance of an Integrative Evolutionary Registry

At this stage, `PhyloCactus` generates an integrative phylogenetic and conservation registry that links evolutionary relationships with complementary biological information, including taxonomic validation, geographic distribution, habitat associations, IUCN conservation assessments, and molecular sampling coverage.

The family **Cactaceae** contains numerous geographically restricted endemics and highly vulnerable evolutionary lineages affected by habitat loss, anthropogenic disturbance, and climate-driven environmental change.

By integrating conservation assessments directly with phylogenetic relationships, this framework enables the identification of **Evolutionarily Distinct and Globally Endangered (EDGE)** lineages, the evaluation of conservation gaps across evolutionary history, and the detection of biases in current molecular sampling efforts.

Taxonomic reconciliation is conducted against the dynamic Cactaceae species checklist published by the Caryophyllales.org initiative (Korotkova *et al*. 2021), comprising 1,965 accepted species.
When combined with 38 outgroup species from Anacampserotaceae, Portulacaceae, and Talinaceae, the total taxonomic evaluation universe encompasses 2,003 taxa.
Across the pipeline, 1,060 species possess sequence data in the multilocus supermatrix (1,022 ingroup and 38 outgroup), while the inferred maximum-likelihood tree and dated chronogram resolve 1,023 terminal tips (985 ingroup and 38 outgroup).
Exactly 37 ingroup species were collapsed by `RAxML-NG` due to identical sequence profiles across their sampled loci.
Crucially, the 943 accepted Cactaceae species lacking sequence data represent critical gaps for biodiversity inventory, among which 134 are categorized as threatened (CR, EN, VU) by the IUCN Red List.

Because reconciling taxonomic identities, molecular datasets, and conservation metadata requires extensive data harmonization, `PhyloCactus` provides dedicated API wrappers (`get_iucn_data(sp_name)`) to facilitate reproducible retrieval and integration of species-level information from the IUCN Red List database.

Now, let's load our libraries and prepare the dataset outputs from previous **Modules**.

```{r, eval = FALSE}
# ============================================================
# LIBRARIES
# ============================================================
library(PhyloCactus)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(ggplot2)
library(forcats)
library(purrr)
library(scales)

dir.create("9_Visualization/figures", showWarnings = FALSE, recursive = TRUE)
dir.create("9_Visualization/tables", showWarnings = FALSE, recursive = TRUE)

# ============================================================
# INPUT FILES (Referenced from tutorial root paths)
# ============================================================
species_file <- "6_Concatenated/final_tables/TABLE_final_species_alignment_summary.csv"
output_file <- "9_Visualization/tables/TABLE_species_all_data.csv"
prioritization_file <- "9_Visualization/tables/TABLE_conservation_prioritization_gaps.csv"
marker_stats_file <- "6_Concatenated/final_tables/TABLE_marker_statistics.csv"
marker_ranges_file <- "6_Concatenated/final_tables/TABLE_marker_ranges.tsv"

# Canonical inputs that do not change across runs
accepted_list_file <- system.file("extdata", "CactaceaeFullList_2026_07_01_Beatriz_Merino.xlsx", package = "PhyloCactus")
constraints_csv_path <- system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus")

# Tree outputs from Stages 9 and 10
supp_tree_file <- if (file.exists("7_Phylogenetics/cactus_support.raxml.support")) {
  "7_Phylogenetics/cactus_support.raxml.support"
} else {
  "7_Phylogenetics/cactus.raxml.bestTree"
}
dated_tree_file <- "8_Dating/BestTree_treePL.tree"
chrono_file <- "8_Dating/dated_summary_hpd.tree"

# ============================================================
# LOAD SPECIES TABLE & ACCEPTED CHECKLIST (Korotkova et al. 2021)
# ============================================================
message("Reading species genetics table from supermatrix outputs...")
species_summary <- read_csv(species_file, show_col_types = FALSE) %>%
  mutate(species_clean = str_replace_all(species, "_", " "))

message("Reading accepted checklist (Korotkova et al. 2021, Caryophyllales.org)...")
if (grepl("\\.xlsx?$", accepted_list_file, ignore.case = TRUE)) {
  sheets <- readxl::excel_sheets(accepted_list_file)
  checklist_sheets <- sheets[!grepl("^facts", sheets, ignore.case = TRUE)]
  list_df <- lapply(checklist_sheets, function(sh) {
    df <- readxl::read_excel(accepted_list_file, sheet = sh)
    df$family_sheet <- sh
    df
  })
  accepted_list <- dplyr::bind_rows(list_df)
} else {
  accepted_list <- read.csv(accepted_list_file, stringsAsFactors = FALSE)
  if (!"family_sheet" %in% names(accepted_list)) accepted_list$family_sheet <- "Cactaceae"
}

# Normalization helper ensuring perfect match with molecular alignment keys
normalize_species <- function(x) {
  x |>
    stringr::str_replace_all("\u00d7", "") |>
    stringr::str_replace_all("\\bx[[:space:]]+", "") |>
    stringr::str_trim() |>
    stringr::str_replace_all("[ -]+", "_")
}

accepted_species_table <- accepted_list %>%
  filter(tolower(family_sheet) == "cactaceae", RANK == "Species") %>%
  mutate(pureName = str_squish(as.character(pureName)),
         fullName = str_squish(as.character(fullName))) %>%
  filter(!str_detect(pureName, regex("\\b(subsp|ssp|var|forma|f\\.|subg|sect|ser|cf\\.|aff\\.|sp\\.|spp\\.|nr\\.)\\b", ignore_case = TRUE))) %>%
  mutate(species = normalize_species(pureName)) %>%
  filter(nzchar(species)) %>%
  select(species, checklist_pureName = pureName, checklist_fullName = fullName,
         checklist_author = author, checklist_rank = RANK, checklist_taxon = taxon, checklist_uuid = uuid) %>%
  distinct(species, .keep_all = TRUE)

message("Reading phylogeny tree tips...")
ml_tree_tips <- ape::read.tree(supp_tree_file)$tip.label
chrono_tips <- if (file.exists(dated_tree_file)) {
  ape::read.tree(dated_tree_file)$tip.label
} else if (file.exists(chrono_file)) {
  as.phylo(treeio::read.beast(chrono_file))$tip.label
} else {
  ml_tree_tips
}

# Identify identical sequence groups collapsed during ML tree inference
raxml_log <- "7_Phylogenetics/cactus.raxml.log"
rx_lines <- if (file.exists(raxml_log)) readLines(raxml_log) else character(0)
warn_lines <- rx_lines[grepl("are exactly identical", rx_lines)]
collapsed_taxa <- if (length(warn_lines) > 0) {
  unique(stringr::str_match(warn_lines, "and ([A-Za-z0-9_]+) are")[, 2])
} else if (file.exists("6_Concatenated/logs_and_qc/SUPP_TABLE_identical_sequence_groups.csv")) {
  read_csv("6_Concatenated/logs_and_qc/SUPP_TABLE_identical_sequence_groups.csv", show_col_types = FALSE)$taxon
} else {
  setdiff(species_summary$species, ml_tree_tips)
}

message("Reconciling taxonomy, sequences, and phylogenetic tips...")
species_summary <- species_summary %>%
  full_join(accepted_species_table, by = "species") %>%
  mutate(
    species_clean = if_else(!is.na(species_clean), species_clean, str_replace_all(species, "_", " ")),
    in_checklist = !is.na(checklist_pureName),
    in_alignment = !is.na(species_class),
    in_ml_tree = species %in% ml_tree_tips,
    in_chrono_tree = species %in% chrono_tips,
    is_outgroup = if_else(in_alignment, species_class == "outgroup", FALSE),
    is_ingroup = if_else(in_alignment, species_class == "ingroup", !is_outgroup),
    # Backward-compatible flags
    has_sequences = in_alignment,
    in_phylogeny = in_ml_tree,
    accepted_in_checklist = in_checklist,
    accepted_or_outgroup = in_checklist | is_outgroup,
    # Categorical pipeline status
    status_pipeline = case_when(
      is_outgroup ~ "outgroup_reference",
      in_checklist & in_alignment & in_chrono_tree ~ "sampled_and_dated",
      in_alignment & !in_ml_tree & (species %in% collapsed_taxa) ~ "sequenced_collapsed_duplicate",
      in_alignment & !in_ml_tree ~ "sequenced_unplaced",
      in_checklist & !in_alignment ~ "unsampled_checklist_gap",
      TRUE ~ "other"
    ),
    record_status = case_when(
      is_outgroup ~ "accepted_outgroup",
      in_ml_tree & in_checklist ~ "sampled_ingroup",
      in_ml_tree & !in_checklist ~ "rejected_ingroup_in_tree",
      in_alignment & in_checklist ~ "sequenced_not_in_tree",
      in_alignment & !in_checklist ~ "rejected_ingroup_no_tree",
      !in_alignment & in_checklist ~ "unsampled_ingroup",
      TRUE ~ "other"
    )
  ) %>%
  distinct(species, .keep_all = TRUE) %>%
  arrange(species_clean)
```

#### IUCN Metadata Enrichment

The integration of conservation information represents an essential step for connecting phylogenetic reconstruction with biological interpretation.
In this module, `PhyloCactus` retrieves species-level conservation assessments from the **IUCN Red List database** through the `rredlist` API interface and integrates these records with the final phylogenetic dataset.

The enrichment process links each taxon included in the molecular framework with standardized conservation attributes, including IUCN category, assessment status, endemic distribution information, habitat associations, and documented threats.
This approach allows researchers to evaluate whether evolutionary diversity is adequately represented within conservation assessments and to identify potential mismatches between phylogenetic distinctiveness and extinction risk.

Because taxonomic discrepancies frequently occur between molecular datasets and external biodiversity databases, the workflow first performs taxonomic reconciliation against the accepted species checklist (Korotkova *et al*. 2021) before retrieving conservation information.
Only validated species-level names are queried against the IUCN API, reducing false matches and ensuring that conservation metadata are associated with the appropriate evolutionary units.

The resulting enriched table (`TABLE_species_all_data.csv`) represents the integrated species registry used throughout the visualization section.
This resource combines phylogenetic placement, molecular sampling information, taxonomic validation, and conservation attributes into a single reproducible dataset for downstream comparative analyses.

```{r, eval = FALSE}
# ============================================================
# RUN IUCN EXTRACTION OR LOAD INTEGRATED SPECIES TABLE
# ============================================================
if (file.exists(output_file)) {
  message("Reading integrated species table: ", output_file)
  species_summary_iucn <- read_csv(output_file, show_col_types = FALSE)
} else {
  message("Extracting IUCN data...")
  iucn_data <- purrr::map_dfr(species_summary$species_clean, PhyloCactus::get_iucn_data)
  species_summary_iucn <- species_summary %>%
    left_join(iucn_data, by = "species_clean")
  write_csv(species_summary_iucn, output_file)
}

# Export conservation prioritization table for unsampled gaps
TABLE_conservation_prioritization_gaps <- species_summary_iucn %>%
  filter(status_pipeline == "unsampled_checklist_gap", iucn_category %in% c("CR", "EN", "VU")) %>%
  mutate(genus = stringr::str_extract(species, "^[A-Za-z]+")) %>%
  select(species, species_clean, genus, iucn_category, iucn_category_name,
         iucn_criteria, iucn_population_trend, iucn_locations, iucn_endemic_locations,
         iucn_is_endemic, checklist_author, checklist_uuid) %>%
  arrange(factor(iucn_category, levels = c("CR", "EN", "VU")), genus, species)

write_csv(TABLE_conservation_prioritization_gaps, prioritization_file)
```

```{r, eval = TRUE, echo = FALSE}
# Check if tutorial root analysis files exist to conditionally evaluate plotting chunks
has_data <- dir.exists(tutorial_dir) &&
  file.exists(file.path(tutorial_dir, "9_Visualization/tables/TABLE_species_all_data.csv")) &&
  file.exists(file.path(tutorial_dir, "6_Concatenated/final_tables/TABLE_marker_statistics.csv"))

if (!has_data) {
  message(
    "Note: Plotting chunks in Module 12 require analytical outputs from earlier pipeline stages.\n",
    "Expected outputs were not found under '", tutorial_dir, "':\n",
    "  - 9_Visualization/tables/TABLE_species_all_data.csv\n",
    "  - 6_Concatenated/final_tables/TABLE_marker_statistics.csv\n",
    "To generate these outputs, execute the pipeline sequentially via Tutorials 1 and 2, ",
    "or run the companion script: inst/scripts/tutorial-3-cactus-phylogeny-visualization.R."
  )
}
```

### Module 12: Comprehensive Data Visualization and Macroevolutionary Summaries

Following the completion of maximum-likelihood inference, divergence time estimation, and biological metadata integration, this module presents the complete analytical synthesis of `PhyloCactus`.

To establish a clear scientific progression, the visualizations are structured into five consecutive analytical parts:

1. **Molecular Dataset and Alignment Properties (Figures 1 to 3)**: Taxonomic marker representation, locus completeness per species, and locus occupancy heatmaps.
2. **Supermatrix Concatenation Structure (Figures 4 to 5)**: Partition architecture, boundary coordinates, and parsimony-informative site contributions.
3. **Inferred Phylogeny and Divergence Timescale (Figures 6 to 9)**: Maximum-likelihood topology with Felsenstein's Bootstrap Proportions (FBP), high-resolution uncollapsed A0 poster, time-calibrated chronogram with 95% HPD intervals, and extended chronogram poster.
4. **Taxonomic Reconciliation and Species Cadastre (Figure 10)**: Global species accounting comparing the taxonomic checklist, supermatrix sampling, collapsed duplicates, and phylogenetic placement.
5. **IUCN Conservation and Macroevolutionary Syntheses (Figures 11 to 17)**: Extinction risk distributions, molecular sampling completeness across threat categories, conservation prioritization of unsampled gaps, national endemism rankings, threat profiles, and ecological habitat associations.

```{r, eval = has_data}
library(PhyloCactus)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(ggplot2)
library(forcats)
library(purrr)
library(scales)

# Harmonized visualization theme
theme_phylocactus <- function(base_size = 12, base_family = "") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      panel.background = element_rect(fill = "#F5F5F5", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "white", linewidth = 0.6),
      panel.grid.minor = element_line(color = "white", linewidth = 0.3),
      axis.ticks = element_line(color = "grey70", linewidth = 0.4),
      axis.title = element_text(face = "italic", color = "grey20"),
      axis.title.x = element_text(face = "italic", margin = margin(t = 8)),
      axis.title.y = element_text(face = "italic", margin = margin(r = 8)),
      axis.text = element_text(color = "grey30"),
      plot.title = element_text(face = "bold", color = "grey15", hjust = 0),
      plot.subtitle = element_text(color = "grey40", hjust = 0, margin = margin(b = 8)),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.key = element_rect(fill = "transparent", color = NA),
      legend.title = element_text(face = "bold", size = rel(0.9)),
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold", color = "grey20", size = rel(0.95))
    )
}

# Load previously written analysis files directly from the tutorial workspace
species_summary_iucn <- read_csv(file.path(tutorial_dir, "9_Visualization/tables/TABLE_species_all_data.csv"), show_col_types = FALSE)

marker_stats <- read_csv(file.path(tutorial_dir, "6_Concatenated/final_tables/TABLE_marker_statistics.csv"), show_col_types = FALSE)
marker_cols <- intersect(marker_stats$marker, names(species_summary_iucn))

prioritization_file <- file.path(tutorial_dir, "9_Visualization/tables/TABLE_conservation_prioritization_gaps.csv")
TABLE_conservation_prioritization_gaps <- if (file.exists(prioritization_file)) {
  read_csv(prioritization_file, show_col_types = FALSE)
} else {
  species_summary_iucn %>%
    filter(status_pipeline == "unsampled_checklist_gap", iucn_category %in% c("CR", "EN", "VU")) %>%
    mutate(genus = stringr::str_extract(species, "^[A-Za-z]+"))
}

# Data Summary Table Generation
TABLE_dataset_summary <- tibble(
  metric = c(
    "Total ingroup species in checklist",
    "Total species with sequences (supermatrix)",
    "Ingroup species with sequences (supermatrix)",
    "Outgroup species with sequences (supermatrix)",
    "Ingroup species sampled and dated in tree",
    "Outgroup reference species in tree",
    "Sequenced ingroup species collapsed duplicates",
    "Unsampled ingroup species (checklist gaps)",
    "Total species in final phylogeny",
    "Ingroup species mapped to IUCN assessment",
    "Ingroup species without IUCN assessment",
    "Threatened species in phylogeny (VU/EN/CR) (Ingroup)",
    "Threatened unsequenced species (VU/EN/CR) (Priority gaps)",
    "Endemic species in phylogeny (Ingroup)"
  ),
  value = c(
    sum(species_summary_iucn$in_checklist & species_summary_iucn$is_ingroup, na.rm = TRUE),
    sum(species_summary_iucn$in_alignment, na.rm = TRUE),
    sum(species_summary_iucn$is_ingroup & species_summary_iucn$in_alignment, na.rm = TRUE),
    sum(species_summary_iucn$is_outgroup & species_summary_iucn$in_alignment, na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "sampled_and_dated", na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "outgroup_reference", na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "sequenced_collapsed_duplicate", na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "unsampled_checklist_gap", na.rm = TRUE),
    sum(species_summary_iucn$in_ml_tree, na.rm = TRUE),
    sum(species_summary_iucn$is_ingroup & species_summary_iucn$iucn_found == TRUE, na.rm = TRUE),
    sum(species_summary_iucn$is_ingroup & (is.na(species_summary_iucn$iucn_found) | species_summary_iucn$iucn_found == FALSE), na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "sampled_and_dated" & species_summary_iucn$iucn_category %in% c("VU", "EN", "CR"), na.rm = TRUE),
    sum(species_summary_iucn$status_pipeline == "unsampled_checklist_gap" & species_summary_iucn$iucn_category %in% c("VU", "EN", "CR"), na.rm = TRUE),
    sum(species_summary_iucn$in_ml_tree & species_summary_iucn$is_ingroup & species_summary_iucn$iucn_is_endemic == TRUE, na.rm = TRUE)
  )
)
knitr::kable(TABLE_dataset_summary)
```

#### Figure 1: Marker Coverage Across Species

This figure summarizes the taxonomic representation of each molecular marker retained in the final dataset by quantifying the number of species with available sequence data across locus partitions.

```{r marker_coverage, eval = has_data}
p1 <- ggplot(marker_stats, aes(x = reorder(marker, n_taxa), y = n_taxa)) +
  geom_col(fill = "grey35", width = 0.7) +
  coord_flip() +
  theme_phylocactus(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(
    title = "Marker Coverage Across Species",
    x = "Marker",
    y = "Number of species"
  )
p1
```

#### Figure 2: Distribution of Marker Completeness

This histogram illustrates the distribution of molecular marker completeness across taxa resolved in the maximum-likelihood tree, indicating the number of retained loci per species following sequence filtering and quality control.

```{r marker_completeness, eval = has_data}
if ("retained_markers" %in% names(species_summary_iucn)) {
  p2 <- ggplot(species_summary_iucn %>% filter(in_ml_tree == TRUE), aes(x = retained_markers)) +
    geom_histogram(binwidth = 1, fill = "grey35", color = "white") +
    scale_x_continuous(breaks = seq(1, 11, 1)) +
    theme_phylocactus(base_size = 12) +
    labs(
      title = "Distribution of Marker Completeness",
      x = "Number of retained markers",
      y = "Number of species"
    )
  p2
}
```

#### Figure 3: Marker Presence Heatmap

This heatmap provides an overview of locus occupancy across all species resolved in the phylogeny, illustrating patterns of data presence and missingness across individual marker partitions.

```{r marker_heatmap, eval = has_data, fig.width = 8, fig.height = 8}
heatmap_data <- species_summary_iucn %>%
  filter(in_ml_tree == TRUE) %>%
  select(species, all_of(marker_cols)) %>%
  pivot_longer(cols = all_of(marker_cols), names_to = "marker", values_to = "accession") %>%
  mutate(present = accession != "-")

p3 <- ggplot(heatmap_data, aes(x = marker, y = fct_rev(species), fill = present)) +
  scale_fill_grey(start = 0.95, end = 0.25, name = "Present", labels = c("No", "Yes")) +
  geom_tile() +
  theme_phylocactus(base_size = 10) +
  labs(
    title = "Marker Presence Heatmap",
    x = "Marker",
    y = "Species"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p3
```

#### Figure 4: Supermatrix Structure

This figure depicts the linear architecture of the concatenated supermatrix, displaying the start and end nucleotide coordinates for each locus partition.

```{r supermatrix_structure, eval = has_data}
ranges_file <- file.path(tutorial_dir, "6_Concatenated/final_tables/TABLE_marker_ranges.tsv")
if (file.exists(ranges_file)) {
  TABLE_marker_ranges <- read_tsv(ranges_file, show_col_types = FALSE)
  if ("marker" %in% names(TABLE_marker_ranges)) {
    TABLE_marker_ranges <- TABLE_marker_ranges %>% mutate(length = end - start + 1)
    p4 <- ggplot(TABLE_marker_ranges) +
      geom_segment(aes(x = start, xend = end, y = marker, yend = marker), linewidth = 6, color = "grey35") +
      theme_phylocactus(base_size = 12) +
      theme(axis.text.y = element_text(face = "italic")) +
      labs(
        title = "Supermatrix Structure",
        x = "Alignment position (bp)",
        y = "Marker"
      )
    p4
  }
}
```

#### Figure 5: Marker Informativeness

We evaluate the phylogenetic utility of each partition by plotting the number of parsimony-informative sites against total alignment length.

```{r marker_informativeness, eval = has_data}
if ("parsimony_informative" %in% names(marker_stats)) {
  p5 <- ggplot(marker_stats, aes(x = alignment_length, y = parsimony_informative, label = marker)) +
    geom_point(size = 3.5, color = "grey35") +
    geom_text(nudge_y = 20, size = 3.2, fontface = "italic", color = "grey20") +
    theme_phylocactus(base_size = 12) +
    labs(
      title = "Marker Informativeness",
      x = "Alignment length (bp)",
      y = "Parsimony informative sites"
    )
  p5
}
```

#### Figure 6: Maximum-Likelihood Tree with Bootstrap Support (FBP)

This plot renders the inferred maximum-likelihood topology annotated with Felsenstein's Bootstrap Proportions (FBP) mapped directly onto internal nodes, alongside major clade demarcation layers.

```{r figure_6_ml_tree, eval = has_data, fig.width = 8.5, fig.height = 11, message = FALSE, warning = FALSE}
library(ggtree)
library(ape)
library(tidytree)

# Load support tree from Stage 9
supp_tree_file <- file.path(tutorial_dir, "7_Phylogenetics/cactus_support.raxml.support")
if (!file.exists(supp_tree_file)) {
  supp_tree_file <- file.path(tutorial_dir, "7_Phylogenetics/cactus.raxml.bestTree")
}

if (file.exists(supp_tree_file) && nzchar(supp_tree_file)) {
  ml_tree <- ape::read.tree(supp_tree_file)
  ml_tree <- ape::ladderize(ml_tree)
  
  # Canonical constraints table for clade annotations
  constraints_csv_path <- system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus")
  if (file.exists(constraints_csv_path) && nzchar(constraints_csv_path)) {
    constraints_tbl <- read_csv(constraints_csv_path, show_col_types = FALSE)
    
    if (is.null(ml_tree$node.label)) {
      ml_tree$node.label <- rep(NA_character_, ml_tree$Nnode)
    }

    MAIN_LEVEL4_VALUES <- c("Anacampseros", "Grahamia", "Talinopsis", "Leuenbergeria", "Pereskia", "Maihuenia", "Blossfeldia", "Opuntieae", "Cylindropuntieae", "Tephrocacteae", "Cacteae", "Core I", "Core II", "Copiapoa", "Calymmanthium", "Rhipsalis", "Portulaca", "Notocacteae", "BCT", "Rhipsalideae")
    ML_MAIN_LAYER_SPECS <- tibble::tribble(
      ~reg_name, ~fontsize, ~barsize, ~offset, ~offset_text, ~fontface, ~sort_desc, ~angle, ~align,
      "level_4_main", 2.5, 0.34, 0.010, 0.0038, "plain", TRUE, 0, TRUE,
      "level_3", 3.5, 0.50, 0.170, 0.0055, "bold", TRUE, 270, TRUE,
      "level_2", 4.0, 0.75, 0.220, 0.0100, "bold", TRUE, 270, TRUE,
      "level_1", 4.5, 0.85, 0.270, 0.0100, "bold", TRUE, 270, TRUE
    )
    
    registry <- PhyloCactus::build_annotation_registry(
      tree = ml_tree,
      constraints_tbl = constraints_tbl,
      main_level4_values = MAIN_LEVEL4_VALUES,
      supp_level4_min_tips = 3L
    )

    max_raw <- max(suppressWarnings(as.numeric(ml_tree$node.label)), na.rm = TRUE)
    is_proportion <- !is.na(max_raw) && max_raw <= 1.0

    # Identify topologically constrained nodes (cactus_constraints.tree)
    # to avoid presenting algorithmic constraints as empirical bootstrap support
    constraint_tree_file <- file.path(tutorial_dir, "7_Phylogenetics/cactus_constraints.tree")
    constrained_nodes_vec <- integer(0)
    if (file.exists(constraint_tree_file) && nzchar(constraint_tree_file)) {
      constrained_df <- tryCatch(
        PhyloCactus::classify_constrained_nodes(ml_tree, constraint_tree_file),
        error = function(e) NULL
      )
      if (!is.null(constrained_df)) {
        constrained_nodes_vec <- constrained_df$node[constrained_df$constrained]
      }
    }

    ml_tree_data <- tidytree::as_tibble(ml_tree) %>%
      mutate(
        raw_val = suppressWarnings(as.numeric(label)),
        pct_val = if (is_proportion) raw_val * 100 else raw_val,
        is_constrained = node %in% constrained_nodes_vec,
        support_label = ifelse(is_constrained | is.na(pct_val), NA_character_, sprintf("%.0f", pct_val)),
        support_class = case_when(
          is_constrained ~ "Constrained",
          pct_val >= 90  ~ ">= 90",
          pct_val >= 70  ~ "70-89",
          TRUE           ~ "< 70"
        ),
        support_class = factor(support_class, levels = c(">= 90", "70-89", "< 70", "Constrained"))
      )
    ml_tree_plot <- tidytree::as.treedata(ml_tree_data)

    p6 <- ggtree(ml_tree_plot, linewidth = 0.3) +
      theme_tree() +
      geom_nodepoint(aes(fill = support_class), shape = 21, size = 2, stroke = 0.2, na.rm = TRUE) +
      scale_fill_manual(
        values = c(">= 90" = "white", "70-89" = "grey", "< 70" = "black", "Constrained" = "skyblue"),
        name = "Support (FBP) / Constraint",
        na.translate = FALSE
      ) +
      theme(legend.position = "bottom")

    p6 <- PhyloCactus::apply_clade_label_layers(p6, registry, ML_MAIN_LAYER_SPECS)
    xmax <- max(p6$data$x, na.rm = TRUE)
    p6 <- p6 + coord_cartesian(xlim = c(0, xmax + xmax * 0.35), clip = "off") + theme(plot.margin = margin(12, 12, 12, 12))

    p6
  }
}
```

#### Figure 7: Extended Constrained ML Phylogeny (A0 Poster)

To examine all 1,023 terminal tips without node compression, this visualization formats the complete topology as an A0-dimension poster, displaying individual species binomials and node-level bootstrap support.

```{r figure_7_ml_tree_extended, eval = FALSE}
if (file.exists(supp_tree_file) && nzchar(supp_tree_file)) {
  ML_SUPP_LAYER_SPECS <- tibble::tribble(
    ~reg_name, ~fontsize, ~barsize, ~offset, ~offset_text, ~fontface, ~sort_desc, ~angle, ~align,
    "level_4_supp", 6, 0.60, 0.030, 0.0038, "plain", TRUE, 0, TRUE,
    "level_3", 8, 0.65, 0.150, 0.0038, "bold", TRUE, 270, TRUE,
    "level_2", 9, 0.70, 0.200, 0.0100, "bold", TRUE, 270, TRUE,
    "level_1", 10, 0.80, 0.250, 0.0100, "bold", TRUE, 270, TRUE
  )
  
  p7_supp <- ggtree(ml_tree_plot, linewidth = 0.2) +
    theme_tree() +
    geom_tiplab(size = 0.70, align = FALSE, linetype = "solid", linesize = 0.15, colour = "grey20") +
    geom_nodelab(aes(label = support_label), size = 0.70, hjust = 0.7, colour = "black", na.rm = TRUE) +
    theme(legend.position = "none")
    
  p7_supp <- PhyloCactus::apply_clade_label_layers(p7_supp, registry, ML_SUPP_LAYER_SPECS)
  xmax_supp <- max(p7_supp$data$x, na.rm = TRUE)
  p7_supp <- p7_supp + coord_cartesian(xlim = c(0, xmax_supp + xmax_supp * 0.30), clip = "off") + theme(plot.margin = ggplot2::margin(12, 12, 12, 12))
  
  ggplot2::ggsave("9_Visualization/figures/Figure_7_ML_Tree_Extended.pdf", plot = p7_supp, width = 36, height = 48, limitsize = FALSE)
}
```

For interactive exploration in external phylogenetic software (such as FigTree), the annotated `treedata` object can be exported into Nexus/BEAST format:

```{r export_beast_ml, eval = FALSE}
annotated_treedata <- PhyloCactus::augment_treedata_with_registry(ml_tree_plot, registry)
treeio::write.beast(annotated_treedata, file = "9_Visualization/figures/annotated_ml_tree.tree")
```

> **Note on Extended Tree Poster & Direct Downloads:**
> Due to its large format dimensions (A0 poster: 36 × 48 inches / 91.4 × 121.9 cm) and 1,023 terminal tips, this extended tree is not rendered inline within this document to maintain browser responsiveness and prevent visual overcrowding.
> 
> When executing this workflow locally, the full vector PDF and annotated BEAST tree file are generated directly in your workspace:
> - Local PDF: `9_Visualization/figures/Figure_7_ML_Tree_Extended.pdf`
> - Local Tree: `9_Visualization/figures/annotated_ml_tree.tree`
> 
> Once the online repository is published on GitHub, these high-resolution assets can be downloaded directly from the repository [`extended_trees/`](https://github.com/beeamerino/PhyloCactus/tree/main/extended_trees) directory:
> - Direct PDF Download: [`Figure_7_ML_Tree_Extended.pdf`](https://raw.githubusercontent.com/beeamerino/PhyloCactus/main/extended_trees/Figure_7_ML_Tree_Extended.pdf)
> - Direct Tree Download: [`annotated_ml_tree.tree`](https://raw.githubusercontent.com/beeamerino/PhyloCactus/main/extended_trees/annotated_ml_tree.tree)

#### Figure 8: Time-Calibrated Chronogram with HPD Intervals

This chronogram displays absolute divergence times estimated via penalized likelihood (`treePL`) and summarized across temporal bootstrap replicates, with node age uncertainty intervals (95% HPD) and a calibrated geological timescale axis.

```{r figure_8_chronogram, eval = has_data, fig.width = 8.5, fig.height = 11, message = FALSE, warning = FALSE}
# Load summary chronogram from Stage 10
chrono_file <- file.path(tutorial_dir, "8_Dating/dated_summary_hpd.tree")
if (!file.exists(chrono_file)) {
  chrono_file <- file.path(tutorial_dir, "8_Dating/BestTree_treePL.tree")
}

if (file.exists(chrono_file) && nzchar(chrono_file)) {
  chrono_beast <- treeio::read.beast(chrono_file)
  
  constraints_csv_path <- system.file("extdata", "cactus_constraints.csv", package = "PhyloCactus")
  if (file.exists(constraints_csv_path) && nzchar(constraints_csv_path)) {
    constraints_tbl <- read_csv(constraints_csv_path, show_col_types = FALSE)
    tree_phy <- as.phylo(chrono_beast)
    
    if (is.null(tree_phy$node.label)) {
      tree_phy$node.label <- rep(NA_character_, tree_phy$Nnode)
    }

    MAIN_LEVEL4_VALUES <- c("Anacampseros", "Grahamia", "Talinopsis", "Leuenbergeria", "Pereskia", "Maihuenia", "Blossfeldia", "Opuntieae", "Cylindropuntieae", "Tephrocacteae", "Cacteae", "Core I", "Core II", "Copiapoa", "Calymmanthium", "Rhipsalis", "Portulaca", "Notocacteae", "BCT", "Rhipsalideae")
    CHRONO_MAIN_LAYER_SPECS <- tibble::tribble(
      ~reg_name, ~fontsize, ~barsize, ~offset, ~offset_text, ~fontface, ~sort_desc, ~angle, ~align,
      "level_4_main", 2.5, 0.34, 0.010, 0.0038, "plain", TRUE, 0, TRUE,
      "level_3", 3.5, 0.50, 0.170, 0.0055, "bold", TRUE, 270, TRUE,
      "level_2", 4.0, 0.75, 0.220, 0.0100, "bold", TRUE, 270, TRUE,
      "level_1", 4.5, 0.85, 0.270, 0.0100, "bold", TRUE, 270, TRUE
    )
    
    registry_chrono <- PhyloCactus::build_annotation_registry(
      tree = tree_phy,
      constraints_tbl = constraints_tbl,
      main_level4_values = MAIN_LEVEL4_VALUES,
      supp_level4_min_tips = 3L
    )

    p8 <- ggtree(chrono_beast, linewidth = 0.3) +
      theme_tree2() + 
      geom_range(range = 'height_0.95_HPD', color = 'gray60', alpha = 0.4, linewidth = 1.5) +
      theme_tree()
    
    p8 <- PhyloCactus::apply_clade_label_layers(p8, registry_chrono, CHRONO_MAIN_LAYER_SPECS)
    p8 <- PhyloCactus::add_chronogram_axis(p8, tree_phy, by = 5, digits = 0L, segment_size = 3, title_margin_top = 30)
    
    xmax <- max(p8$data$x, na.rm = TRUE)
    p8 <- p8 + coord_cartesian(xlim = c(0, xmax + xmax * 0.35), clip = "off") + theme(plot.margin = margin(12, 12, 12, 12))

    p8
  }
}
```

#### Figure 9: Extended Time-Calibrated Chronogram (A0 Poster)

This A0-dimension poster presents the uncollapsed temporal chronogram with all 1,023 tip names, divergence time confidence intervals, and geological period boundaries.

```{r figure_9_chronogram_extended, eval = FALSE}
if (file.exists(chrono_file) && nzchar(chrono_file)) {
  CHRONO_SUPP_LAYER_SPECS <- tibble::tribble(
    ~reg_name, ~fontsize, ~barsize, ~offset, ~offset_text, ~fontface, ~sort_desc, ~angle, ~align,
    "level_4_supp", 6, 0.60, 0.030, 0.0038, "plain", TRUE, 0, TRUE,
    "level_3", 8, 0.65, 0.150, 0.0038, "bold", TRUE, 270, TRUE,
    "level_2", 9, 0.70, 0.200, 0.0100, "bold", TRUE, 270, TRUE,
    "level_1", 10, 0.80, 0.250, 0.0100, "bold", TRUE, 270, TRUE
  )
  
  p9_supp <- ggtree(chrono_beast, linewidth = 0.2) +
    theme_tree2() +
    geom_tiplab(size = 0.85, align = FALSE, linetype = "solid", linesize = 0.15, colour = "grey20") +
    geom_range(range = 'height_0.95_HPD', color = 'gray60', alpha = 0.4, linewidth = 0.8) +
    theme_tree()
    
  p9_supp <- PhyloCactus::apply_clade_label_layers(p9_supp, registry_chrono, CHRONO_SUPP_LAYER_SPECS)
  p9_supp <- PhyloCactus::add_chronogram_axis(p9_supp, tree_phy, by = 5, digits = 0L, segment_size = 10, title_margin_top = 28, bar_size = 30)
  xmax_chrono_supp <- max(p9_supp$data$x, na.rm = TRUE)
  p9_supp <- p9_supp + coord_cartesian(xlim = c(0, xmax_chrono_supp + xmax_chrono_supp * 0.30), clip = "off") + theme(plot.margin = ggplot2::margin(12, 12, 32, 12))
  
  ggplot2::ggsave("9_Visualization/figures/Figure_9_Chronogram_Extended.pdf", plot = p9_supp, width = 36, height = 48, limitsize = FALSE)
}
```

For interactive exploration in external phylogenetic software (such as FigTree), the annotated `treedata` object can be exported into Nexus/BEAST format:

```{r export_beast_chrono, eval = FALSE}
annotated_chrono <- PhyloCactus::augment_treedata_with_registry(chrono_beast, registry_chrono)
treeio::write.beast(annotated_chrono, file = "9_Visualization/figures/annotated_chronogram.tree")
```

> **Note on Extended Chronogram Poster & Direct Downloads:**
> Due to its large format dimensions (A0 poster: 36 × 48 inches / 91.4 × 121.9 cm), 1,023 terminal tips, and node age uncertainty intervals (95% HPD), this extended chronogram is not rendered inline within this document to maintain browser responsiveness.
> 
> When executing this workflow locally, the full vector PDF and annotated BEAST chronogram file are generated directly in your workspace:
> - Local PDF: `9_Visualization/figures/Figure_9_Chronogram_Extended.pdf`
> - Local Tree: `9_Visualization/figures/annotated_chronogram.tree`
> 
> Once the online repository is published on GitHub, these high-resolution assets can be downloaded directly from the repository [`extended_trees/`](https://github.com/beeamerino/PhyloCactus/tree/main/extended_trees) directory:
> - Direct PDF Download: [`Figure_9_Chronogram_Extended.pdf`](https://raw.githubusercontent.com/beeamerino/PhyloCactus/main/extended_trees/Figure_9_Chronogram_Extended.pdf)
> - Direct Tree Download: [`annotated_chronogram.tree`](https://raw.githubusercontent.com/beeamerino/PhyloCactus/main/extended_trees/annotated_chronogram.tree)

#### Figure 10: Species Composition (Phylogeny vs. Checklist by Pipeline Status)

This figure summarizes the methodological composition of the taxonomic universe across pipeline stages, distinguishing taxa that were sampled and dated in the tree (985), sequenced identical duplicates collapsed by `RAxML-NG` (37), unsampled checklist gaps (943), and outgroup reference species (38).

```{r figure_10_species_composition, eval = has_data}
p10 <- ggplot(species_summary_iucn %>% count(status_pipeline), aes(x = reorder(status_pipeline, n), y = n)) +
  geom_col(fill = "grey35", width = 0.7) +
  coord_flip() +
  geom_text(aes(label = scales::comma(n)), hjust = -0.15, size = 3.8, color = "grey20") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_phylocactus(base_size = 12) +
  labs(
    title = "Species Composition (Phylogeny vs Checklist)",
    subtitle = "Methodological status across checklist, supermatrix, and tree",
    x = NULL,
    y = "Number of species"
  )
p10
```

#### Figure 11: IUCN Categories in the Ingroup Phylogeny

This bar chart summarizes the distribution of IUCN Red List conservation categories among the ingroup species resolved in the final phylogenetic reconstruction.

```{r figure_11_iucn_categories, eval = has_data}
if (any(species_summary_iucn$iucn_found, na.rm = TRUE)) {
  TABLE_iucn_categories <- species_summary_iucn %>%
    filter(iucn_found == TRUE, in_ml_tree == TRUE, is_ingroup == TRUE) %>%
    count(iucn_category, iucn_category_name, sort = TRUE)
  
  p11 <- ggplot(TABLE_iucn_categories, aes(x = reorder(iucn_category, n), y = n, fill = iucn_category)) +
    PhyloCactus::scale_fill_iucn(name = "IUCN Category") +
    geom_col(width = 0.7) +
    theme_phylocactus(base_size = 12) +
    labs(
      title = "IUCN Categories (Ingroup Phylogeny)",
      x = "Category",
      y = "Number of species"
    )
  p11
}
```

#### Figure 12: Molecular Completeness vs. IUCN Category by Pipeline Status

This plot compares molecular marker completeness against IUCN threat categories, faceted by pipeline status (`sampled_and_dated` vs `sequenced_collapsed_duplicate`), to test whether extinction risk correlates with missing molecular data or sequence collapse during phylogenetic inference.

```{r figure_12_completeness_vs_iucn, eval = has_data, fig.width = 8.5}
if (any(species_summary_iucn$iucn_found, na.rm = TRUE)) {
  p12 <- ggplot(
    species_summary_iucn %>%
      filter(status_pipeline %in% c("sampled_and_dated", "sequenced_collapsed_duplicate"),
             iucn_found == TRUE,
             !is.na(iucn_category)),
    aes(x = iucn_category, y = pct_markers, fill = iucn_category)
  ) +
    geom_violin(trim = FALSE, alpha = 0.5) +
    geom_boxplot(width = 0.2, outlier.size = 0.6, alpha = 0.8) +
    PhyloCactus::scale_fill_iucn(name = "IUCN Category") +
    facet_wrap(~ status_pipeline, labeller = as_labeller(c(
      "sampled_and_dated" = "Sampled & Dated in Tree (N = 985)",
      "sequenced_collapsed_duplicate" = "Sequenced Collapsed Duplicates (N = 37)"
    ))) +
    theme_phylocactus(base_size = 12) +
    labs(
      title = "Molecular Completeness vs IUCN Category by Pipeline Status",
      subtitle = "Comparison of marker occupancy between retained tips and collapsed identical sequences",
      x = "IUCN Category",
      y = "Marker completeness (%)"
    )
  p12
}
```

#### Figure 13: Conservation Prioritization of Unsampled Gaps by Genus

This figure visualizes the taxonomic distribution of threatened species (CR, EN, VU) among the 943 checklist gaps that currently lack molecular sequences in public repositories.
By highlighting genera with the highest concentrations of unsequenced threatened taxa (led by *Parodia*, *Mammillaria*, and *Melocactus*), this analysis establishes explicit priorities for targeted field collecting and sequencing.

```{r figure_13_conservation_prioritization_gaps, eval = has_data, fig.height = 5}
plot_gaps_genus <- TABLE_conservation_prioritization_gaps %>%
  count(genus, iucn_category) %>%
  mutate(iucn_category = factor(iucn_category, levels = c("CR", "EN", "VU")))

top_gap_genera <- TABLE_conservation_prioritization_gaps %>%
  count(genus, sort = TRUE) %>%
  slice_max(n, n = 12) %>%
  pull(genus)

p13 <- ggplot(plot_gaps_genus %>% filter(genus %in% top_gap_genera),
              aes(x = factor(genus, levels = rev(top_gap_genera)), y = n, fill = iucn_category)) +
  geom_col(width = 0.7) +
  coord_flip() +
  PhyloCactus::scale_fill_iucn(name = "Threat Category") +
  theme_phylocactus(base_size = 12) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(
    title = "Threatened Unsampled Species by Genus (Sampling Gaps)",
    subtitle = "Priority Cactaceae taxa (CR, EN, VU) lacking molecular sequences",
    x = "Genus",
    y = "Number of threatened unsequenced species"
  )
p13
```

#### Figure 14: Endemic Species by Country

This figure summarizes national cactus endemism represented in the phylogenetic tree, ranking countries by the number of endemic species resolved in the molecular framework.

```{r figure_14_endemism_by_country, eval = has_data}
if (any(species_summary_iucn$iucn_is_endemic, na.rm = TRUE)) {
  plot_data_endemism <- species_summary_iucn %>%
    filter(iucn_is_endemic == TRUE, in_ml_tree == TRUE, is_ingroup == TRUE) %>%
    separate_rows(iucn_endemic_locations, sep = "; ") %>%
    count(iucn_endemic_locations, sort = TRUE) %>%
    slice_max(n, n = 20)
    
  p14 <- ggplot(plot_data_endemism, aes(x = reorder(iucn_endemic_locations, n), y = n)) +
    geom_col(fill = "grey35", width = 0.7) +
    coord_flip() +
    theme_phylocactus(base_size = 12) +
    labs(
      title = "Endemic Species by Country (Ingroup Phylogeny)",
      x = "Country",
      y = "Number of endemic species"
    )
  p14
}
```

#### Figure 15: Endemic Species by Country and IUCN Category

This stacked bar chart visualizes the conservation status of national endemics across the top cactus-diverse countries, breaking down endemic richness by IUCN threat category.

```{r figure_15_iucn_by_country, eval = has_data}
if (any(species_summary_iucn$iucn_found, na.rm = TRUE)) {
  plot_iucn_country <- species_summary_iucn %>%
    filter(iucn_is_endemic == TRUE, in_ml_tree == TRUE, is_ingroup == TRUE, !is.na(iucn_endemic_locations)) %>%
    separate_rows(iucn_endemic_locations, sep = "; ") %>%
    count(iucn_endemic_locations, iucn_category)
    
  top_countries <- plot_data_endemism$iucn_endemic_locations
  plot_iucn_country <- plot_iucn_country %>% filter(iucn_endemic_locations %in% top_countries)
  
  if (nrow(plot_iucn_country) > 0) {
    p15 <- ggplot(plot_iucn_country, aes(x = factor(iucn_endemic_locations, levels = rev(top_countries)), y = n, fill = iucn_category)) +
      geom_col(width = 0.7) +
      coord_flip() +
      PhyloCactus::scale_fill_iucn(name = "IUCN Category") +
      theme_phylocactus(base_size = 12) +
      labs(
        title = "Endemic Species by Country and IUCN Category",
        x = "Country",
        y = "Number of endemic species"
      )
    p15
  }
}
```

#### Figure 16: Primary Ecological Threats

This bar chart illustrates the most frequently documented environmental and anthropogenic threats impacting cactus species in the ingroup phylogeny, based on IUCN Red List threat classifications.

```{r figure_16_threats, eval = has_data}
if (any(!is.na(species_summary_iucn$iucn_threat_names))) {
  plot_threats <- species_summary_iucn %>%
    filter(!is.na(iucn_threat_names), in_ml_tree == TRUE, is_ingroup == TRUE) %>%
    separate_rows(iucn_threat_names, sep = "; ") %>%
    count(iucn_threat_names, sort = TRUE) %>%
    slice_max(n, n = 15)
    
  if (nrow(plot_threats) > 0) {
    p16 <- ggplot(plot_threats, aes(x = reorder(iucn_threat_names, n), y = n)) +
      geom_col(fill = "grey35", width = 0.7) +
      coord_flip() +
      theme_phylocactus(base_size = 12) +
      labs(
        title = "Most Common Threats (Ingroup Phylogeny)",
        x = "Threat",
        y = "Number of species"
      )
    p16
  }
}
```

#### Figure 17: Primary Ecological Habitats

This bar chart displays the primary habitat associations recorded for species in the ingroup phylogeny, identifying major ecological biomes supporting cactus diversification.

```{r figure_17_habitats, eval = has_data}
if (any(!is.na(species_summary_iucn$iucn_habitat_names))) {
  plot_habitats <- species_summary_iucn %>%
    filter(!is.na(iucn_habitat_names), in_ml_tree == TRUE, is_ingroup == TRUE) %>%
    separate_rows(iucn_habitat_names, sep = "; ") %>%
    count(iucn_habitat_names, sort = TRUE) %>%
    slice_max(n, n = 15)
    
  if (nrow(plot_habitats) > 0) {
    p17 <- ggplot(plot_habitats, aes(x = reorder(iucn_habitat_names, n), y = n)) +
      geom_col(fill = "grey35", width = 0.7) +
      coord_flip() +
      theme_phylocactus(base_size = 12) +
      labs(
        title = "Most Common Habitats (Ingroup Phylogeny)",
        x = "Habitat",
        y = "Number of species"
      )
    p17
  }
}
```

## Conclusion

At this stage, phylogenetic visualization and conservation metadata integration have been successfully completed.
The resulting outputs provide a comprehensive representation of the evolutionary framework generated by `PhyloCactus`, including the inferred maximum-likelihood topology, the time-calibrated chronogram, molecular sampling completeness, and associated conservation information.

The next stage focuses on **phylogenetic validation and comparative analyses**, where the robustness of the focal phylogenetic hypothesis will be evaluated against previously published evolutionary frameworks and alternative topological hypotheses.

Using quantitative tree comparison approaches, including **Robinson-Foulds distances** and **multidimensional scaling (MDS)**, the following tutorial will assess the degree of topological agreement among alternative phylogenetic hypotheses, identify regions of phylogenetic uncertainty, and evaluate how differences in taxon sampling and analytical strategies influence the recovered evolutionary relationships.

[Continue to Tutorial 4: Phylogenetic Validation and Comparative Analyses](https://beeamerino.github.io/PhyloCactus/articles/tutorial-4-cactus-phylogeny-validation.html)

## References

- Amaral *et al*. 2022. Spatial patterns of evolutionary diversity in Cactaceae show low ecological representation within protected areas. *Biological Conservation*, 273, 109677. <https://doi.org/10.1016/j.biocon.2022.109677>
- Korotkova *et al*. 2021. Cactaceae at Caryophyllales.org - A dynamic online species-level taxonomic backbone for the family. *Willdenowia*, 51(2), 251–270. <https://doi.org/10.3372/wi.51.51208>
- Thompson *et al*. 2024. Identifying the multiple drivers of cactus diversification. *Nature Communications*, 15(1), 7114. <https://doi.org/10.1038/s41467-024-51666-2>
